---
title: "Untitled"
output: html_document
date: "2025-08-27"
---

Load libraries.

```{r}
library(dplyr)
library(ggplot2)
library(readxl)
library(ggpubr)
library(rstatix)
library(multcomp)
```

Read in data.

```{r}
ymaze_data <- read_excel("Female cohort - PO/Data to plot - female cohort.xlsx", sheet = "Ymaze")
ymaze_data$Value <- as.numeric(ymaze_data$Value)
ymaze_data$`Mouse Group` <- ifelse(ymaze_data$`Mouse Group` == 'Veh (PO)', 'Veh\n(PO)', 'G_50\n(PO)')
ymaze_data$`Mouse Group` <- factor(ymaze_data$`Mouse Group`,
                                levels = c("Veh\n(PO)", "G_50\n(PO)"))
```

Function to plot % alternation data.

According to Nay's email - "In the Ymaze, you don’t compare between the groups. You have to run a 1 sample t test comparing the data to 50% chance level."

```{r}
plotting_func_alternation <- function(df, variable, ylab, fname, wid = 2, ht = 2) {
  df <- dplyr::filter(df, Measure == variable)
  
  # Compute statistics
  # One sample t-test vs. chance (50% alternation) per group
  # Have to remove NA data first
  df <- dplyr::filter(df, !is.na(Value))
  tests <- df %>%
    dplyr::group_by(`Mouse Group`) %>%
    rstatix::t_test(Value ~ 1, mu = 50, detailed = TRUE) %>%
    rstatix::add_significance() %>%
    dplyr::ungroup()
  
  # Define positions of significance stars
  sig_pos <- df %>%
    dplyr::group_by(`Mouse Group`) %>%
    dplyr::summarise(y.position = min(100, max(Value, na.rm = TRUE) + 5), .groups = "drop")
  
  # Merge
  tests <- merge(tests, sig_pos, by = "Mouse Group") %>%
    dplyr::mutate(label = p.signif)
  
  ggplot(df, aes(x=`Mouse Group`, y=Value)) +
    geom_boxplot(aes(color = `Mouse Group`), fill = NA, outlier.shape = NA, linewidth = 1) +
    geom_point(color = "grey", position = position_jitter(width = 0.1), size = 1) +
    scale_color_manual(values = c("Veh\n(PO)" = "#E38AB8", "G_50\n(PO)" = "#E0B83E")) +
    geom_text(data = tests, aes(x = `Mouse Group`, y = y.position, label = label), vjust = 0, size = 3.5) +
    theme_bw() +
    theme(legend.position = "none",
      axis.line = element_line(color='black'),
      axis.text.y = element_text(size = 8, color = 'black'),
      axis.text.x = element_text(size = 10, color = 'black'),
      axis.title.y = element_text(size = 10, color = 'black'),
      axis.title.x = element_blank()) +
    labs(y = ylab) +
    geom_hline(yintercept = 50, linetype = 'dashed', color = 'black') +
    ylim(20,100)
  
  ggsave(fname, width = wid, height = ht)
  
  return(tests)
}
```

Apply the plotting function. 

```{r}
# "% alternation (0-5min)"
pc_alt_05 <- plotting_func_alternation(ymaze_data, "% alternation (0-5min)",
              "% alternation (0-5min)", "Female cohort - PO/ymaze_pc_alternation_0_to_5.pdf")
```

Plot line graph for entries over time.

```{r}
entries <- dplyr::filter(ymaze_data, startsWith(Measure, 'Entries'))

# Add a timepoint column
entries$Time <- substr(entries$Measure, 12, 12)

# Need a "summary" data frame
entries_summary <- entries %>%
  group_by(`Mouse Group`, Time) %>%
  summarise(
    mean_value = mean(Value, na.rm = TRUE),
    se = sd(Value, na.rm = TRUE) / sqrt(n()),
    .groups = "drop"
  )

# Want to get rid of the newline for the line graph legend
entries_summary$`Mouse Group` <- ifelse(entries_summary$`Mouse Group` == 'Veh\n(PO)', 'Veh (PO)', 'G_50 (PO)')
entries_summary$`Mouse Group` <- factor(entries_summary$`Mouse Group`,
                                        levels = c("Veh (PO)", "G_50 (PO)"))

entries_summary <- dplyr::rename(entries_summary, `Female Mouse\nGroup`=`Mouse Group`)

# Plot line graph
 
ggplot(entries_summary, aes(x = Time, y = mean_value, color = `Female Mouse\nGroup`, group = `Female Mouse\nGroup`)) + geom_line() + geom_point() + geom_errorbar(aes(ymin = mean_value - se, ymax = mean_value + se), width = 0.2) + labs(x = "Time (Min)", y = "Cumulative arm entries") +
    scale_color_manual(values = c("Veh (PO)" = "#E38AB8", "G_50 (PO)" = "#E0B83E")) +
    theme_bw() +
    theme(axis.line = element_line(color='black'),
      axis.text = element_text(size = 8, color = 'black'),
      axis.title = element_text(size = 10, color = 'black'),
      legend.title = element_text(size = 10, color = 'black'),
      legend.text = element_text(size = 8, color = 'black'))

ggsave('Female cohort - PO/ymaze_cumulative_arm_entries.pdf', width = 4, height = 2)
```

Also want to make a boxplot for entries (0-5min).

```{r}
generic_boxplot_func <- function(df, variable, ylab,
                          fname, ymax, wid = 2, ht = 2) {
  
  df <- dplyr::filter(df, Measure == variable)
  ggplot(df, aes(x=`Mouse Group`, y=Value)) +
    geom_boxplot(aes(color = `Mouse Group`), fill = NA, outlier.shape = NA, linewidth = 1) +
    geom_point(color = "grey", position = position_jitter(width = 0.1), size = 1) +
    scale_color_manual(values = c("Veh\n(PO)" = "#E38AB8", "G_50\n(PO)" = "#E0B83E")) +
    theme_bw() +
    theme(legend.position = "none",
      axis.line = element_line(color='black'),
      axis.text.y = element_text(size = 8, color = 'black'),
      axis.text.x = element_text(size = 10, color = 'black'),
      axis.title.y = element_text(size = 10, color = 'black'),
      axis.title.x = element_blank()) +
    labs(y = ylab) + ylim(0, ymax)
  
  ggsave(fname, width = wid, height = ht)
}
```

Apply the function and print out the stats.

```{r}
generic_boxplot_func(entries, 'Entries (0-5min)', 'Entries (0-5min)',
                     'Female cohort - PO/ymaze_entries_05_boxplot.pdf', 60)

for_stats <- dplyr::filter(entries, Measure == 'Entries (0-5min)')
t.test(dplyr::filter(for_stats, `Mouse Group` == 'Veh\n(PO)')$Value,
       dplyr::filter(for_stats, `Mouse Group` == 'G_50\n(PO)')$Value)
```

Plot line graph for distance moved over time.

```{r}
distance <- dplyr::filter(ymaze_data, startsWith(Measure, 'Distance'))

# Add a timepoint column
distance$Time <- rep(seq(1,5), each = 30)

# Need a "summary" data frame
distance_summary <- distance %>%
  group_by(`Mouse Group`, Time) %>%
  summarise(
    mean_value = mean(Value, na.rm = TRUE),
    se = sd(Value, na.rm = TRUE) / sqrt(n()),
    .groups = "drop"
  )

# Want to get rid of the newline for the line graph legend
distance_summary$`Mouse Group` <- ifelse(distance_summary$`Mouse Group` == 'Veh\n(PO)', 'Veh (PO)', 'G_50 (PO)')
distance_summary$`Mouse Group` <- factor(distance_summary$`Mouse Group`, levels = c("Veh (PO)", "G_50 (PO)"))

distance_summary <- dplyr::rename(distance_summary, `Female Mouse\nGroup`=`Mouse Group`)

# Plot line graph
ggplot(distance_summary, aes(x = Time, y = mean_value, color = `Female Mouse\nGroup`, group = `Female Mouse\nGroup`)) + geom_line() + geom_point() + geom_errorbar(aes(ymin = mean_value - se, ymax = mean_value + se), width = 0.2) + labs(x = "Time (Min)", y = "Distance moved (cm/min)") +
    scale_color_manual(values = c("Veh (PO)" = "#E38AB8", "G_50 (PO)" = "#E0B83E")) +
    theme_bw() +
    theme(axis.line = element_line(color='black'),
      axis.text = element_text(size = 8, color = 'black'),
      axis.title = element_text(size = 10, color = 'black'),
      legend.title = element_text(size = 10, color = 'black'),
      legend.text = element_text(size = 8, color = 'black'))

ggsave('Female cohort - PO/ymaze_distance_moved_per_min.pdf', width = 4, height = 2)
```

Also make boxplot for cumulative distance moved by the end, and print out the stats.

```{r}
distance_sum <- dplyr::filter(ymaze_data, Measure == "Total distance moved (cm)")

generic_boxplot_func(distance_sum, 'Total distance moved (cm)', 'Total distance (cm)', 'Female cohort - PO/ymaze_total_distance_boxplot.pdf', 2000)

t.test(dplyr::filter(distance_sum, `Mouse Group` == 'Veh\n(PO)')$Value,
       dplyr::filter(distance_sum, `Mouse Group` == 'G_50\n(PO)')$Value)
```
