---
title: "Untitled"
output: html_document
date: "2025-08-19"
---

I will consult this helpful tutorial: https://biostatsquid.com/fgsea-tutorial-gsea/

```{r}
library("fgsea")
library(dplyr)
library(ggplot2)

# Set seed for reproducibility
set.seed(123)
```

Prepare our list of background genes for TMAO.

```{r}
tmao_deseq2_res <- read.csv("deseq2_outputs/tmao_all_results.txt", sep = '\t')
# Filter to genes with gene names and DESeq padj not NA (padj NA occurs sometimes in certain situations - https://bioconductor.org/packages/release/bioc/vignettes/DESeq2/inst/doc/DESeq2.html#pvaluesNA)
tmao_deseq2_res <- dplyr::filter(tmao_deseq2_res, Gene_Name != '', !is.na(padj))
# Define background genes
tmao_bg <- tmao_deseq2_res$Gene_Name
```

Downloaded .gmt files from: https://www.gsea-msigdb.org/gsea/msigdb/mouse/collections.jsp

Specifically, downloaded:

MH: hallmark gene sets
Reactome subset of CP (canonical pathways)

Define function to prepare GMT file for fgsea.

```{r}
prepare_gmt <- function(gmt_file, bg_genes) {
  # Read in file
  gmt <- gmtPathways(gmt_file)
  all_genes <- unique(unlist(gmt))
  
  # Convert gmt file to a matrix with genes as rows and values are 0 or 1 per pathway
  mat <- matrix(NA, dimnames = list(all_genes, names(gmt)), nrow = length(all_genes), ncol = length(gmt))
  for (i in 1:dim(mat)[2]){
    mat[,i] <- as.numeric(all_genes %in% gmt[[i]])
  }
  
  # Subset to genes in our bg_genes list
  genes_in_bg_genes <- intersect(bg_genes, all_genes)
  mat <- mat[genes_in_bg_genes,]
  
  # Need to convert to list
  pathways_list <- list()
  for (pathway in colnames(mat)) {
    pathways_list[[pathway]] <- rownames(mat)[as.logical(mat[, pathway])]
  }
  
  return(pathways_list)
}
```

Apply the function - prepare the GMTs!

```{r}
mh_gmt <- prepare_gmt("mh.all.v2025.1.Mm.symbols.gmt", tmao_bg)
reactome_gmt <- prepare_gmt("m2.cp.reactome.v2025.1.Mm.symbols.gmt", tmao_bg)
```

Prepare ranked gene list.

To rank the genes, will use the ranking metric from the blog post above.

"Here, I use the sign(df$log2fc)*(-log10(df$pval))) as a ranking metric.

This way, the magnitude of the ranking will be given by the p value (since I use -log10(), the more significant a gene is differentially expressed, the larger that value will be). The sign of the ranking (positive or negative) will be given by the fold-change: upregulated genes will have positive rankings, downregulated genes will have negative rankings."

Also supported by: https://www.nature.com/articles/s41596-018-0103-9
https://doi.org/10.1038/s41596-018-0103-9
"The gene list is then ranked by one or more of these values (e.g., −log10 P value multiplied by the sign of log-transformed fold-change)"

```{r}
tmao_deseq2_res$rankings <- sign(tmao_deseq2_res$log2FoldChange) * (-log10(tmao_deseq2_res$pvalue))

# Realized there are some duplicated gene names --> get rid of those extra rows. Keep the row with the highest ranking
tmao_deseq2_res <- tmao_deseq2_res %>%
  group_by(Gene_Name) %>%
  slice_max(order_by = abs(rankings), n = 1) %>%
  ungroup()

rankings_for_fgsea <- tmao_deseq2_res$rankings
names(rankings_for_fgsea) <- tmao_deseq2_res$Gene_Name
```

fgsea function.

```{r}
fgsea_func <- function(prepared_gmt, ranked_genes,
                       fname) {
  result <- fgsea(pathways = prepared_gmt, stats = ranked_genes, minSize = 10, maxSize = 500)
  
  # Write out results
  results <- as.data.frame(result)
  results$leadingEdge <- sapply(results$leadingEdge, paste, collapse=",")
  # order by absolute value of NES
  results <- results[order(abs(results$NES), decreasing = TRUE), ]
  write.table(results, fname, sep = '\t',
              quote = FALSE, row.names = FALSE)
  
  return(result)
}
```

Apply the function for each of our prepared GMTs!

```{r}
# MH
mh_res_tmao <- fgsea_func(mh_gmt, rankings_for_fgsea,
                     'fgsea_results/mh_results_tmao.txt')

# Reactome
reactome_res_tmao <- fgsea_func(reactome_gmt, rankings_for_fgsea,
                      'fgsea_results/reactome_results_tmao.txt')
```

Now, do the same for GDCA!

```{r}
# Prepare background genes
gdca_deseq2_res <- read.csv("deseq2_outputs/gdca_all_results.txt", sep = '\t')
gdca_deseq2_res <- dplyr::filter(gdca_deseq2_res, Gene_Name != '', !is.na(padj))
gdca_bg <- gdca_deseq2_res$Gene_Name

# Prepare GMTs
mh_gmt <- prepare_gmt("mh.all.v2025.1.Mm.symbols.gmt", gdca_bg)
reactome_gmt <- prepare_gmt("m2.cp.reactome.v2025.1.Mm.symbols.gmt", gdca_bg)

# Rankings
gdca_deseq2_res$rankings <- sign(gdca_deseq2_res$log2FoldChange) * (-log10(gdca_deseq2_res$pvalue))

gdca_deseq2_res <- gdca_deseq2_res %>%
  group_by(Gene_Name) %>%
  slice_max(order_by = abs(rankings), n = 1) %>%
  ungroup()

rankings_for_fgsea <- gdca_deseq2_res$rankings
names(rankings_for_fgsea) <- gdca_deseq2_res$Gene_Name

# MH
mh_res_gdca <- fgsea_func(mh_gmt, rankings_for_fgsea,
                     'fgsea_results/mh_results_gdca.txt')

# Reactome
reactome_res_gdca <- fgsea_func(reactome_gmt, rankings_for_fgsea,
                      'fgsea_results/reactome_results_gdca.txt')
```

Make plots!

First, define a helpful function.

```{r}
# from https://www.r-bloggers.com/2019/11/setting-axes-to-integer-values-in-ggplot2-2/
# A function factory for getting integer y-axis values.
integer_breaks <- function(n = 5, ...) {
  fxn <- function(x) {
    breaks <- floor(pretty(x, n, ...))
    names(breaks) <- attr(breaks, "labels")
    breaks
  }
  return(fxn)
}
```

First plot - TMAO, MH results of interest.

```{r}
mh_res_tmao <- read.csv('fgsea_results/mh_results_tmao.txt', sep = '\t')
mh_res_tmao <- dplyr::filter(mh_res_tmao, pathway %in% c("HALLMARK_TNFA_SIGNALING_VIA_NFKB", "HALLMARK_CHOLESTEROL_HOMEOSTASIS", "HALLMARK_REACTIVE_OXYGEN_SPECIES_PATHWAY", "HALLMARK_P53_PATHWAY", "HALLMARK_HYPOXIA", "HALLMARK_MTORC1_SIGNALING", "HALLMARK_OXIDATIVE_PHOSPHORYLATION", "HALLMARK_GLYCOLYSIS"))

# Sanity check
stopifnot(max(mh_res_tmao$padj) <= 0.05)

# eliminate hallmark prefix
mh_res_tmao$pathway <- substr(mh_res_tmao$pathway, 10, nchar(mh_res_tmao$pathway))

# Convert to sentence case
mh_res_tmao <- mh_res_tmao %>%
  mutate(
    pathway = case_when(
      pathway == "TNFA_SIGNALING_VIA_NFKB" ~ "TNFA signaling via NFKB",
      pathway == "CHOLESTEROL_HOMEOSTASIS" ~ "Cholesterol homeostasis",
      pathway == "P53_PATHWAY" ~ "p53 pathway",
      pathway == "REACTIVE_OXYGEN_SPECIES_PATHWAY" ~ "Reactive oxygen species pathway",
      pathway == "HYPOXIA" ~ "Hypoxia",
      pathway == "MTORC1_SIGNALING" ~ "mTORC1 signaling",
      pathway == "OXIDATIVE_PHOSPHORYLATION" ~ "Oxidative phosphorylation",
      pathway == "GLYCOLYSIS" ~ "Glycolysis"
    )
  )

# Add a neglog10padj column
mh_res_tmao$neglog10padj <- -log10(mh_res_tmao$padj)

# put in order
mh_res_tmao <- mh_res_tmao[order(abs(mh_res_tmao$NES), decreasing = TRUE), ]
mh_res_tmao$pathway <- factor(mh_res_tmao$pathway,
                             levels = mh_res_tmao$pathway)

ggplot(mh_res_tmao, aes(x=pathway, y=NES, fill=neglog10padj)) +
  geom_bar(stat="identity") +
  scale_x_discrete(limits = rev(levels(mh_res_tmao$pathway))) + coord_flip() +
  scale_fill_gradient(label = function(x) format(x, digits=3),
                      low = "#b2b2d8", high = "#000080") +
  theme_bw() + theme(panel.border = element_blank(), 
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(), 
      axis.line = element_line(colour = "black"),
      axis.text.x=element_text(size=8, color = "black"), 
      axis.text.y=element_text(size=10, color='black'),
      axis.title.x=element_text(size=10, color='black'),
      axis.title=element_blank(),
      legend.title=element_text(size=8, color='black'),
      legend.text=element_text(size=8, color='black')) +
  labs(y = 'Normalized Enrichment Score (NES)', fill = "-log10(padj)") +
scale_y_continuous(breaks = integer_breaks()) 

ggsave('fgsea_results/fgsea_mh_tmao.pdf', width = 4, height = 3)
```

Next - TMAO, REACTOME results of interest.

```{r}
reactome_res_tmao <- read.csv('fgsea_results/reactome_results_tmao.txt', sep = '\t')
reactome_res_tmao <- dplyr::filter(reactome_res_tmao, pathway %in% c("REACTOME_CHOLESTEROL_BIOSYNTHESIS", "REACTOME_GABA_SYNTHESIS_RELEASE_REUPTAKE_AND_DEGRADATION", "REACTOME_INTEGRATION_OF_ENERGY_METABOLISM", "REACTOME_MRNA_SPLICING", "REACTOME_GABA_B_RECEPTOR_ACTIVATION", "REACTOME_MAPK_FAMILY_SIGNALING_CASCADES", "REACTOME_NEURONAL_SYSTEM", "REACTOME_SIGNALING_BY_NTRKS"))

# Sanity check
stopifnot(max(reactome_res_tmao$padj) <= 0.05)

# eliminate prefix
reactome_res_tmao$pathway <- substr(reactome_res_tmao$pathway, 10, nchar(reactome_res_tmao$pathway))

# change pathways to sentence case
reactome_res_tmao <- reactome_res_tmao %>%
  mutate(pathway = case_when(
   pathway == "CHOLESTEROL_BIOSYNTHESIS" ~ "Cholesterol biosynthesis",
   pathway == "GABA_SYNTHESIS_RELEASE_REUPTAKE_AND_DEGRADATION" ~ "GABA synthesis, release,\nreuptake and degradation",
   pathway == "INTEGRATION_OF_ENERGY_METABOLISM" ~ "Integration of energy metabolism",
   pathway == "GABA_B_RECEPTOR_ACTIVATION" ~ "GABA_B receptor activation",
   pathway == "MRNA_SPLICING" ~ "mRNA splicing",
   pathway == "SIGNALING_BY_NTRKS" ~ "Signaling by NTRKs",
   pathway == "MAPK_FAMILY_SIGNALING_CASCADES" ~ "MAPK family signaling cascades",
   pathway == "NEURONAL_SYSTEM" ~ "Neuronal system"
  ))

# Add a neglog10padj column
reactome_res_tmao$neglog10padj <- -log10(reactome_res_tmao$padj)

reactome_res_tmao <- reactome_res_tmao[order(reactome_res_tmao$NES, decreasing = TRUE),]

reactome_res_tmao$pathway <- factor(reactome_res_tmao$pathway, levels = reactome_res_tmao$pathway)

ggplot(reactome_res_tmao, aes(x=pathway, y=NES, fill=neglog10padj)) +
  geom_bar(stat="identity") +
  scale_x_discrete(limits = rev(levels(reactome_res_tmao$pathway))) + coord_flip() +
  scale_fill_gradient(label = function(x) format(x, digits=3),
                      low = "#b2b2d8", high = "#000080") +
  theme_bw() + theme(panel.border = element_blank(), 
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(), 
      axis.line = element_line(colour = "black"),
      axis.text.x=element_text(size=8, color = "black"), 
      axis.text.y=element_text(size=10, color='black'),
      axis.title.x=element_text(size=10, color='black'),
      axis.title=element_blank(),
      legend.title=element_text(size=8, color='black'),
      legend.text=element_text(size=8, color='black')) +
  labs(y = 'Normalized Enrichment Score (NES)', fill = "-log10(padj)") + scale_y_continuous(breaks = integer_breaks()) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "#008888", linewidth = 1) 

ggsave('fgsea_results/fgsea_reactome_tmao.pdf', width = 4, height = 3)
```

Next, significant GDCA MH results.

```{r}
mh_res_gdca <- read.csv('fgsea_results/mh_results_gdca.txt', sep = '\t')
mh_res_gdca <- dplyr::filter(mh_res_gdca,
                             padj <= 0.05)
# eliminate hallmark prefix
mh_res_gdca$pathway <- substr(mh_res_gdca$pathway, 10, nchar(mh_res_gdca$pathway))

# change pathway to sentence case
mh_res_gdca <- mh_res_gdca %>% mutate(
  pathway = case_when(
    pathway == "TNFA_SIGNALING_VIA_NFKB" ~ "TNFA signaling via NFKB",
    pathway == "HYPOXIA" ~ "Hypoxia",
    pathway == "EPITHELIAL_MESENCHYMAL_TRANSITION" ~ "Epithelial mesenchymal transition",
    pathway == "G2M_CHECKPOINT" ~ "G2M checkpoint"
  )
)

# Add a neglog10padj column
mh_res_gdca$neglog10padj <- -log10(mh_res_gdca$padj)
# order by decreasing NES
mh_res_gdca <- mh_res_gdca[order(mh_res_gdca$NES, decreasing = FALSE),]

# Set order
mh_res_gdca$pathway <- factor(mh_res_gdca$pathway,
                              levels = mh_res_gdca$pathway)

# Plot
ggplot(mh_res_gdca, aes(x=pathway, y=NES, fill=neglog10padj)) +
  geom_bar(stat="identity") +
  scale_x_discrete(limits = rev(levels(mh_res_gdca$pathway))) + coord_flip() +
  scale_fill_gradient(label = function(x) format(x, digits=3),
                      low = "#b2b2d8", high = "#000080") +
  theme_bw() + theme(panel.border = element_blank(), 
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(), 
      axis.line = element_line(colour = "black"),
      axis.text.x=element_text(size=8, color = "black"), 
      axis.text.y=element_text(size=10, color='black'),
      axis.title.x=element_text(size=10, color='black'),
      axis.title=element_blank(),
      legend.title=element_text(size=8, color='black'),
      legend.text=element_text(size=8, color='black')) +
  labs(y = 'Normalized Enrichment Score (NES)', fill = "-log10(padj)") +
scale_y_continuous(breaks = integer_breaks()) + geom_hline(yintercept = 0, linetype = "dashed", color = "#008888", linewidth = 1) 

ggsave('fgsea_results/fgsea_mh_gdca.pdf', width = 4, height = 2.5) 
```

Next, significant GDCA REACTOME results.

```{r}
reactome_res_gdca <- read.csv('fgsea_results/reactome_results_gdca.txt', sep = '\t')
reactome_res_gdca <- dplyr::filter(reactome_res_gdca,
                             padj <= 0.05)
# eliminate prefix
reactome_res_gdca$pathway <- substr(reactome_res_gdca$pathway, 10, nchar(reactome_res_gdca$pathway))

# Change pathways to sentence case
reactome_res_gdca <- reactome_res_gdca %>% mutate(
  pathway = case_when(
    pathway == "PKMTS_METHYLATE_HISTONE_LYSINES" ~
      "PKMTs methylate histone lysines",
    pathway == "PROCESSING_OF_CAPPED_INTRON_CONTAINING_PRE_MRNA" ~
      "Processing of capped intron-\ncontaining pre-mRNA",
    pathway == "EPIGENETIC_REGULATION_OF_GENE_EXPRESSION" ~
      "Epigenetic regulation\nof gene expression",
    pathway == "MRNA_SPLICING" ~ "mRNA splicing",
    pathway == "CHROMATIN_ORGANIZATION" ~ "Chromatin organization"
  )
)

# Add a neglog10padj column
reactome_res_gdca$neglog10padj <- -log10(reactome_res_gdca$padj)
# order by decreasing NES
reactome_res_gdca <- reactome_res_gdca[order(reactome_res_gdca$NES, decreasing = TRUE),]

# Set order
reactome_res_gdca$pathway <- factor(reactome_res_gdca$pathway, levels = reactome_res_gdca$pathway)

# Plot
ggplot(reactome_res_gdca, aes(x=pathway, y=NES, fill=neglog10padj)) +
  geom_bar(stat="identity") +
  scale_x_discrete(limits = rev(levels(reactome_res_gdca$pathway))) + coord_flip() +
  scale_fill_gradient(label = function(x) format(x, digits=3),
                      low = "#b2b2d8", high = "#000080") +
  theme_bw() + theme(panel.border = element_blank(), 
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(), 
      axis.line = element_line(colour = "black"),
      axis.text.x=element_text(size=8, color = "black"), 
      axis.text.y=element_text(size=10, color='black'),
      axis.title.x=element_text(size=10, color='black'),
      axis.title=element_blank(),
      legend.title=element_text(size=8, color='black'),
      legend.text=element_text(size=8, color='black')) +
  labs(y = 'Normalized Enrichment Score (NES)', fill = "-log10(padj)") +
scale_y_continuous(breaks = integer_breaks())

ggsave('fgsea_results/fgsea_reactome_gdca.pdf', width = 4, height = 2.5) 
```