---
title: "Untitled"
output: html_document
date: "2025-08-19"
---

Load libraries.

```{r}
library("dplyr")
library("tximport")
library(ggplot2)
library("DESeq2")
library("rhdf5")
library("apeglm")
library("readxl")
library(stringr)
library(reshape2)
library(ggpubr)
library(ggrepel)
```
Read in the data frame that came with the kallisto index, which links ensembl transcripts to ensembl genes.

```{r}
t2g <- read.csv('t2g.txt', sep = '\t', header = FALSE)
# Rename columns
t2g <- t2g %>%
  dplyr::rename(
    Transcript_ID = V1,
    Gene_ID = V2,
    Gene_Name = V3,
    Transcript_Name = V4,
    Chromosome = V5,
    Start_Position = V6,
    End_Position = V7,
    Strand_Direction = V8
  )

# will need this later when outputting DEGs
t2g_no_transcripts <- dplyr::select(t2g, Gene_ID, Gene_Name) %>% distinct()

# make a minimal version of t2g
t2g <- dplyr::select(t2g, Transcript_ID, Gene_ID)
```

Read in the kallisto-generated estimated counts using tximport.

https://bioconductor.org/packages/devel/bioc/vignettes/tximport/inst/doc/tximport.html#kallisto_with_TSV_files

First, write a tximport function that can read in the kallisto transcript-level counts for samples of interest.

```{r}
tximport_function <- function(sample_condition_df, kallisto_dir) {
  # Construct file paths for each sample’s Kallisto output
  files <- file.path(kallisto_dir, paste0('kallisto_out_', sub("^[^_]*_", "", sample_condition_df$sample)), paste0(sample_condition_df$sample, "_abundance.tsv"))
  names(files) <- sample_condition_df$sample
  
  # Read in the corresponding kallisto results and aggregate to gene level
  txi.kallisto <- tximport(files, type = "kallisto", tx2gene = t2g, txIn = TRUE, txOut = FALSE)
  # explicitly specifying TRUE/FALSE for txIn and txOut even though they're the default according to the tximport manual
  # tx2gene needs transcript id as column 1, gene id as column 2
  
  return(txi.kallisto)
}
```

Now, actually apply it!

```{r}
# Read in conditions table
conditions <- read_excel('sample_info_mouse_brain_rnaseq.xlsx')

# Make two "minimal" versions of this conditions table
# One for Experiment 1, one for Experiment 2
conditions_minimal <- dplyr::select(conditions, Experiment,
                                    `Sample name`, Treatment)
colnames(conditions_minimal) <- c("Experiment", "sample", "condition")
tmao_conditions_minimal <- dplyr::filter(conditions_minimal, Experiment == 1) %>%
  dplyr::select(., -Experiment)
expt2_conditions_minimal <- dplyr::filter(conditions_minimal, Experiment == 2) %>%
  dplyr::select(., -Experiment)

tmao_conditions_minimal$condition <- factor(tmao_conditions_minimal$condition, levels = c("Saline", "TMAO"))

expt2_conditions_minimal$condition <- factor(expt2_conditions_minimal$condition, levels = c("Saline", "GDCA", "Saline + DMSO", "PA", "IAA"))

# Apply tximport function
txi_tmao_samples <- tximport_function(tmao_conditions_minimal, '/path/to/expt1/kallisto/files')
txi_expt2_samples <- tximport_function(expt2_conditions_minimal, '/path/to/expt2/kallisto/files')
```

Want to make a PCA plot for each of these DFs. Use the vst option from DESeq2.

First - DESeq2 function.

```{r}
deseq2_function <- function(sample_condition_df, txi, smallestGroupSize=3, a=0.10) {
  # Apparently absolutely critical for analysis that the column names
  # of the counts table from txi matches the order of samples in sample_condition_df
  # (see here - https://www.bioconductor.org/packages/release/bioc/vignettes/DESeq2/inst/doc/DESeq2.html#count-matrix-input)
  sample_condition_df <- tibble::column_to_rownames(sample_condition_df, var = 'sample')
  stopifnot(all(colnames(txi$counts) == rownames(sample_condition_df)))
  
  # Create the DESeq2 dataset
  # code from https://www.bioconductor.org/packages/release/bioc/vignettes/DESeq2/inst/doc/DESeq2.html#tximport
  dds <- DESeqDataSetFromTximport(txi, colData = sample_condition_df, design = ~condition)
  
  # Pre-filter low-count genes
  # https://www.bioconductor.org/packages/release/bioc/vignettes/DESeq2/inst/doc/DESeq2.html#pre-filtering
  keep <- rowSums(counts(dds) >= 10) >= smallestGroupSize
  dds <- dds[keep,]
  
  # Run DESeq2 analysis, return results with specified alpha level
  # https://www.bioconductor.org/packages/release/bioc/vignettes/DESeq2/inst/doc/DESeq2.html#quick-start
  dds <- DESeq(dds)
  # Extract the condition effect by name
  cond_name <- grep("^condition_", resultsNames(dds), value = TRUE)
  print(cond_name)
  dds_results <- results(dds, name = cond_name, alpha = a)
  # Shrink coefficients
  dds_results <- lfcShrink(dds, coef=cond_name, type = "apeglm", res = dds_results)
  
  # Return results and DESeq2 object
  return(list(results = dds_results, dds = dds))
}
```

Now - PCAs.

https://bioconductor.org/packages/devel/bioc/vignettes/DESeq2/inst/doc/DESeq2.html#extracting-transformed-values 

https://bioconductor.org/packages/devel/bioc/vignettes/DESeq2/inst/doc/DESeq2.html#principal-component-plot-of-the-samples

Want blind = TRUE because we're going for a non-biased PCA

```{r}
# TMAO samples
deseq2_output_tmao_samples <- deseq2_function(tmao_conditions_minimal, txi_tmao_samples, smallestGroupSize=4)

# variance stabilizing transformation
vsd <- vst(deseq2_output_tmao_samples[[2]], blind = TRUE)
pca_res <- plotPCA(vsd, returnData = TRUE)
percentVar <- round(100 * attr(pca_res, "percentVar"))

# Want shorter name for label, and add an underscore
pca_res$name <- pca_res$name %>%
  str_extract("[^_]+$") %>%
  str_replace("([A-Za-z]+)([0-9]+)$", "\\1_\\2")

ggplot(pca_res, aes(x=PC1, y=PC2, color=as.factor(condition), label=name)) + geom_point() + geom_text(nudge_x = 0.3, nudge_y = 0.3, show.legend = FALSE) +
    labs(x = paste0('PC1 (', percentVar[1], '% Variance Explained)'),
         y = paste0('PC2 (', percentVar[2], '% Variance Explained)'),
         color = 'Condition') + theme_bw() + 
  theme(panel.border = element_blank(), 
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(), 
        axis.line = element_line(colour = "black"), 
        axis.title=element_text(size=14, color='black'), 
        axis.text.x=element_text(size=12, color = "black"), 
        axis.text.y=element_text(size=12, color='black'),
        legend.position = "top",
        legend.text=element_text(size=12, color = 'black'),
        legend.title=element_blank()) 
  
ggsave('expt1_samples_pca.pdf', width = 4, height = 5)

# All other samples
expt2_conditions_minimal_dummy <- expt2_conditions_minimal
expt2_conditions_minimal_dummy$condition <- factor(c(rep('Dummy1', 10), rep('Dummy2', 10)), levels = c('Dummy1', 'Dummy2'))

deseq2_output_expt2_samples <- deseq2_function(expt2_conditions_minimal_dummy, txi_expt2_samples, smallestGroupSize=10)

vsd <- vst(deseq2_output_expt2_samples[[2]], blind = TRUE)
pca_res <- plotPCA(vsd, returnData = TRUE)
percentVar <- round(100 * attr(pca_res, "percentVar"))

# Want to color by batch for final figure
pca_res$batch <- paste0("Batch",
                        substr(pca_res$name, nchar(pca_res$name), nchar(pca_res$name)))

# Want shorter name for label, and add an underscore
pca_res$name <- pca_res$name %>%
  str_extract("[^_]+$") %>%
  str_replace("([A-Za-z]+)([0-9]+)$", "\\1_\\2")

ggplot(pca_res, aes(x=PC1, y=PC2, color=as.factor(batch), label=name)) + geom_point() + geom_text(nudge_x = 0.3, nudge_y = 0.3, show.legend = FALSE) +
    labs(x = paste0('PC1 (', percentVar[1], '% Variance Explained)'),
         y = paste0('PC2 (', percentVar[2], '% Variance Explained)'),
         color = 'Condition') + theme_bw() + 
  theme(panel.border = element_blank(), 
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(), 
        axis.line = element_line(colour = "black"), 
        axis.title=element_text(size=14, color='black'), 
        axis.text.x=element_text(size=12, color = "black"), 
        axis.text.y=element_text(size=12, color='black'),
        legend.position = "top",
        legend.text=element_text(size=12, color = 'black'),
        legend.title=element_blank()) 
  
ggsave('expt2_samples_pca.pdf', width = 4, height = 5)
```

Conduct DESeq2 for TMAO outlier sample vs. other samples.

```{r}
tmao_conditions_minimal_outlier <- tmao_conditions_minimal
tmao_conditions_minimal_outlier$condition <- ifelse(tmao_conditions_minimal_outlier$sample == 'Expt1_TMAO3', 'Outlier', 'Not Outlier')
tmao_conditions_minimal_outlier$condition <- factor(tmao_conditions_minimal_outlier$condition, levels = c("Not Outlier", "Outlier"))
deseq2_output_tmao_outlier_vs_others <- deseq2_function(tmao_conditions_minimal_outlier, txi_tmao_samples, smallestGroupSize=1)
```

Function to output DESeq2 results.

Output multiple versions.
1) no changes, just filter to the significant results (padj <= 0.1 and abs(LFC) >= 0.1)
2) "" + filter to LFC > 0 (upregulated genes)
3) "" + filter to LFC < 0 (downregulated genes)

Also output a file with *all* the results, not just the significant ones.

```{r}
output_degs <- function(dds_results, outdir, outfprefix, alpha=0.10) {
  dds_results <- tibble::rownames_to_column(data.frame(dds_results), var = 'Gene_ID')
  
  dds_results <- merge(dds_results, t2g_no_transcripts, by = 'Gene_ID', all.x=TRUE, all.y=FALSE) # this adds the external gene names
  
  # ensure the relevant columns are numeric
  dds_results$padj <- as.numeric(dds_results$padj)
  dds_results$log2FoldChange <- as.numeric(dds_results$log2FoldChange)
  
  # add a column without the gene version
  dds_results$Gene_ID_truncated <- sub("\\..*", "", dds_results$Gene_ID)
  
  # filter to the significant results
  sig <- dplyr::filter(dds_results, padj <= alpha, abs(log2FoldChange) >= 0.1)
  
  # Version 1
  outfname_1 <- paste0(outdir, outfprefix, '_sig.txt')
  write.table(sig, outfname_1, sep='\t', row.names=FALSE, quote=FALSE)
  
  # Version 2
  outfname_2 <- paste0(outdir, outfprefix, '_sig_UP.txt')
  write.table(dplyr::filter(sig, log2FoldChange > 0),
              outfname_2, sep='\t', row.names=FALSE, quote=FALSE)
  
  # Version 3
  outfname_3 <- paste0(outdir, outfprefix, '_sig_DOWN.txt')
  write.table(dplyr::filter(sig, log2FoldChange < 0),
              outfname_3, sep='\t', row.names=FALSE, quote=FALSE)
  
  # Version 4
  outfname_4 <- paste0(outdir, outfprefix, '_all_results.txt')
  write.table(dds_results, outfname_4, sep='\t', row.names=FALSE, quote=FALSE)
}
```

Output DEGs (and all DESeq2 results) for TMAO outlier sample vs. others.

```{r}
output_degs(deseq2_output_tmao_outlier_vs_others[[1]],
            'deseq2_outputs/', 'tmao_outlier_vs_others')
```

Run DESeq2 - TMAO vs. saline samples - including the outlier - to generate the Cook's distance plot mentioned in the methods.

```{r}
deseq2_output_tmao_with_outlier <- deseq2_function(tmao_conditions_minimal, txi_tmao_samples, smallestGroupSize=4)

dds_with_cooks <- deseq2_output_tmao_with_outlier[[2]]
cooks <- data.frame(assays(dds_with_cooks)[["cooks"]])

# Summarize mean and SEM per sample
sample_cooks_df <- data.frame(
  sample = colnames(cooks),
  mean_cooks = apply(cooks, 2, mean, na.rm = TRUE),
  sd_cooks = apply(cooks, 2, sd, na.rm = TRUE),
  n_genes = apply(cooks, 2, function(x) sum(!is.na(x)))
) %>%
  mutate(
    sem_cooks = sd_cooks / sqrt(n_genes)
  ) %>%
  arrange(desc(mean_cooks))

# Simpler name for sample
sample_cooks_df$sample <- sample_cooks_df$sample %>%
  str_extract("[^_]+$") %>%
  str_replace("([A-Za-z]+)([0-9]+)$", "\\1_\\2")

sample_cooks_df$sample <- factor(sample_cooks_df$sample, levels = sample_cooks_df$sample)

ggplot(sample_cooks_df, aes(x = sample, y = mean_cooks)) +
  geom_col(width = 0.7, fill = "#008888") +
  geom_errorbar(
    aes(
      ymin = mean_cooks - sem_cooks,
      ymax = mean_cooks + sem_cooks
    ),
    width = 0.2
  ) +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, color = "black", size = 12),
    axis.text.y = element_text(color = "black", size = 12),
    axis.title.y = element_text(color = "black", size = 14)
  ) +
  labs(
    x = NULL,
    y = "Mean Cook's distance ± SEM",
    title = "Mean Cook's distance\nper sample"
  )

ggsave("cooks_distance_tmao_samples.pdf", width = 3, height = 5)
```

Make the plot, for the supplement, showing the abundance of hemoglobin genes across samples.

```{r}
hb_genes <- data.frame(gene_names = c("Hba-a1", "Hba-a2", "Hbb-bt", "Hbb-bs"),
                       gene_ids = c("ENSMUSG00000069919.8",
                                    "ENSMUSG00000069917.8",
                                    "ENSMUSG00000073940.4",
                                    "ENSMUSG00000052305.7"))

tmao_hb_abundances <- data.frame(txi_tmao_samples$abundance) %>%
  tibble::rownames_to_column(., var = "Gene") %>%
  dplyr::filter(., Gene %in% hb_genes$gene_ids)

tmao_hb_abundances <- merge(tmao_hb_abundances, hb_genes, by.x = "Gene", by.y = "gene_ids")
tmao_hb_abundances <- dplyr::select(tmao_hb_abundances, -Gene)

expt2_hb_abundances <- data.frame(txi_expt2_samples$abundance) %>%
  tibble::rownames_to_column(., var = "Gene") %>%
  dplyr::filter(., Gene %in% hb_genes$gene_ids)

expt2_hb_abundances <- merge(expt2_hb_abundances, hb_genes, by.x = "Gene", by.y = "gene_ids")
expt2_hb_abundances <- dplyr::select(expt2_hb_abundances, -Gene)

# Merge
tmao_hb_abundances <- dplyr::rename(tmao_hb_abundances,
                                    S_1_Expt1=Expt1_S1,
                                    S_2_Expt1=Expt1_S2,
                                    S_3_Expt1=Expt1_S3,
                                    S_4_Expt1=Expt1_S3,
                                    TMAO_1 = Expt1_TMAO1,
                                    TMAO_2 = Expt1_TMAO2,
                                    TMAO_3 = Expt1_TMAO2,
                                    TMAO_4 = Expt1_TMAO2)

expt2_hb_abundances <- dplyr::rename(expt2_hb_abundances,
                                     S_1_Expt2=Expt2_S1,
                                     S_2_Expt2=Expt2_S2,
                                     S_3_Expt2=Expt2_S3,
                                     S_4_Expt2=Expt2_S4,
                                     GDCA_1=Expt2_GDCA1,
                                     GDCA_2=Expt2_GDCA2,
                                     GDCA_3=Expt2_GDCA3,
                                     GDCA_4=Expt2_GDCA4,
                                     IAA_1=Expt2_IAA1,
                                     IAA_2=Expt2_IAA2,
                                     IAA_3=Expt2_IAA3,
                                     IAA_4=Expt2_IAA4,
                                     PA_1=Expt2_PA1,
                                     PA_2=Expt2_PA2,
                                     PA_3=Expt2_PA3,
                                     PA_4=Expt2_PA4,
                                     SD_1=Expt2_SD1,
                                     SD_2=Expt2_SD2,
                                     SD_3=Expt2_SD3,
                                     SD_4=Expt2_SD4)

merged_hb_abundances <- merge(tmao_hb_abundances, expt2_hb_abundances,
                              by = "gene_names")

merged_hb_abundances$gene_names <- factor(merged_hb_abundances$gene_names,
                                          levels = c("Hba-a1", "Hbb-bt", "Hba-a2", "Hbb-bs"))

long_hb <- reshape2::melt(
  merged_hb_abundances,
  id.vars = "gene_names"
)

ggplot(long_hb, aes(x = variable, y = value, group = gene_names)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2) +
  facet_wrap(~ gene_names, ncol = 1, scales = "free_y") +
  theme_classic() +
  labs(
    x = "Sample",
    y = "TPM") +
  theme(
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, color = "black", size = 10),
    axis.text.y = element_text(color = "black", size = 10),
    axis.title.x = element_blank(),
    axis.title.y = element_text(color = "black", size = 14),
    strip.text = element_text(color = "black", size = 14),
    legend.position = "none")

ggsave("hemoglobin_genes_plot.pdf", width = 5, height = 5)
```

Now run DESeq2 for TMAO excluding the outlier sample.

```{r}
# Fix conditions table
tmao_conditions_minimal <- dplyr::filter(tmao_conditions_minimal,
                                         sample != 'Expt1_TMAO3')
tmao_conditions_minimal$condition <- factor(tmao_conditions_minimal$condition,
                                            levels = c("Saline", "TMAO"))
# Run tximport again
txi_tmao_samples <- tximport_function(tmao_conditions_minimal, '/path/to/expt1/kallisto/files')
# DESeq2
deseq2_output_tmao <- deseq2_function(tmao_conditions_minimal, txi_tmao_samples, smallestGroupSize=3)
# Output DESeq2 results
output_degs(deseq2_output_tmao[[1]],
            'deseq2_outputs/', 'tmao')
```

Volcano plot!

```{r}
tmao_res <- read.csv('deseq2_outputs/tmao_all_results.txt', sep='\t')

# Filter out the genes with padj NA since they won't be shown in the volcano plot anyway
tmao_res <- dplyr::filter(tmao_res, !is.na(padj))

# Label a few genes
tmao_res$Label <- ifelse(tmao_res$Gene_Name %in% c('Mir124a-1hg', 'Tia1', 'Arl4d', 'Leng8', 'Slc2a1', 'Srrm2', 'Eml5', 'Atp13a3', 'Per1', 'Amy1', 'Pnisr', 'Sacs', 'Tbce', 'Rbm39', 'Jun', 'Bfsp2', 'Dusp1'), tmao_res$Gene_Name, NA)

# Define color categories
tmao_res$color_category <- ifelse(tmao_res$padj <= 0.10 & tmao_res$log2FoldChange >= 0.1, "Upregulated", 
                      ifelse(tmao_res$padj <= 0.10 & tmao_res$log2FoldChange <= -0.1, "Downregulated", "n.s."))

# Volcano plot
ggplot(tmao_res, aes(x = log2FoldChange, y = -log10(padj))) +
  geom_point(aes(color = color_category), alpha = 0.6, size = 1.5) +
  scale_color_manual(values = c("Upregulated" = "#9f329f", "Downregulated" = "#008888", "n.s." = "grey")) +
  theme_bw() +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray70", linewidth = 0.5) +
    geom_hline(yintercept = 1, linetype = "dashed", color = "gray70", linewidth = 0.5) +
  theme(legend.position = "none",
        axis.text = element_text(size = 12, color = 'black'),
        axis.title = element_text(size = 14, color = 'black'),
        plot.title = element_text(size = 14, hjust = 0.5, color = 'black')) +
  labs(x = "Log2 Fold Change vs. Control",
       y = "-Log10 adjusted p-value",
       title = "Volcano plot for TMAO (1.8 mg/kg)") +
      geom_text_repel(aes(label = Label),
                    na.rm = TRUE,
                    size = 3,
                    color = 'black',
                    segment.size = 0.4,
                    segment.color = "black",
                    force = 1.5,
                    direction = "both",
                    box.padding = 0.5,
                    point.padding = 0.3,
                    min.segment.length = 0.1,
                    max.overlaps = Inf)

ggsave('volcano_plot_tmao.pdf', width = 4, height = 4)
```

Now move on to experiment 2 samples! First, apply tximport.

```{r}
# PA
palmitate_conditions_minimal <- dplyr::filter(expt2_conditions_minimal, condition %in% c("Saline + DMSO", "PA"))
palmitate_conditions_minimal$condition <- ifelse(as.character(palmitate_conditions_minimal$condition) == "Saline + DMSO", "Saline_DMSO", as.character(palmitate_conditions_minimal$condition))
palmitate_conditions_minimal$condition <- factor(palmitate_conditions_minimal$condition, levels = c("Saline_DMSO", "PA"))
txi_pa_samples <- tximport_function(palmitate_conditions_minimal, '/path/to/expt2/kallisto/files')

# IAA
iaa_conditions_minimal <- dplyr::filter(expt2_conditions_minimal, condition %in% c("Saline + DMSO", "IAA"))
iaa_conditions_minimal$condition <- ifelse(as.character(iaa_conditions_minimal$condition) == "Saline + DMSO", "Saline_DMSO", as.character(iaa_conditions_minimal$condition))
iaa_conditions_minimal$condition <- factor(iaa_conditions_minimal$condition, levels = c("Saline_DMSO", "IAA"))
txi_iaa_samples <- tximport_function(iaa_conditions_minimal, '/path/to/expt2/kallisto/files')

# GDCA
gdca_conditions_minimal <- dplyr::filter(expt2_conditions_minimal, condition %in% c("Saline", "GDCA"))
gdca_conditions_minimal$condition <- factor(gdca_conditions_minimal$condition, levels = c("Saline", "GDCA"))
txi_gdca_samples <- tximport_function(gdca_conditions_minimal, '/path/to/expt2/kallisto/files')
```

For DESeq2, will need to incorporate batch information.

```{r}
deseq2_function_with_batch <- function(sample_condition_df, txi, smallestGroupSize=4, a=0.10) {
  # Apparently absolutely critical for analysis that the column names
  # of the counts table from txi matches the order of samples in sample_condition_df
  # (see here - https://www.bioconductor.org/packages/release/bioc/vignettes/DESeq2/inst/doc/DESeq2.html#count-matrix-input)
  sample_condition_df <- tibble::column_to_rownames(sample_condition_df, var = 'sample')
  stopifnot(all(colnames(txi$counts) == rownames(sample_condition_df)))
  
  # Create the DESeq2 dataset
  # code from https://www.bioconductor.org/packages/release/bioc/vignettes/DESeq2/inst/doc/DESeq2.html#tximport
  # Also here
  # https://bioconductor.org/packages/devel/bioc/vignettes/tximport/inst/doc/tximport.html#Downstream_DGE_in_Bioconductor
  dds <- DESeqDataSetFromTximport(txi, colData = sample_condition_df,
                                  design  = ~ batch + condition)
  
  # Pre-filter low-count genes
  # https://www.bioconductor.org/packages/release/bioc/vignettes/DESeq2/inst/doc/DESeq2.html#pre-filtering
  keep <- rowSums(counts(dds) >= 10) >= smallestGroupSize
  dds <- dds[keep,]
  
  # Run DESeq2 analysis, return results with specified alpha level
  # https://www.bioconductor.org/packages/release/bioc/vignettes/DESeq2/inst/doc/DESeq2.html#quick-start
  dds <- DESeq(dds)
  # Extract the condition effect by name
  cond_name <- grep("^condition_", resultsNames(dds), value = TRUE)
  dds_results <- results(dds, name = cond_name, alpha = a)
  # Shrink coefficients
  dds_results <- lfcShrink(dds, coef=cond_name, type = "apeglm", res=dds_results)
  
  print(cond_name)
  
  # Return results and DESeq2 object
  return(list(results = dds_results, dds = dds))
}
```

Incorporate batch info into the relevant conditions DFs.

```{r}
palmitate_conditions_minimal$batch <- paste0("Batch", substr(palmitate_conditions_minimal$sample, nchar(palmitate_conditions_minimal$sample), nchar(palmitate_conditions_minimal$sample)))

iaa_conditions_minimal$batch <- paste0("Batch", substr(iaa_conditions_minimal$sample, nchar(iaa_conditions_minimal$sample), nchar(iaa_conditions_minimal$sample)))

gdca_conditions_minimal$batch <- paste0("Batch", substr(gdca_conditions_minimal$sample, nchar(gdca_conditions_minimal$sample), nchar(gdca_conditions_minimal$sample)))
```

Apply the function and output DESeq2 results!

```{r}
deseq2_output_pa <- deseq2_function_with_batch(palmitate_conditions_minimal, txi_pa_samples)
output_degs(deseq2_output_pa[[1]],
            'deseq2_outputs/', 'pa')

deseq2_output_iaa <- deseq2_function_with_batch(iaa_conditions_minimal, txi_iaa_samples)
output_degs(deseq2_output_iaa[[1]],
            'deseq2_outputs/', 'iaa')

deseq2_output_gdca <- deseq2_function_with_batch(gdca_conditions_minimal, txi_gdca_samples)
output_degs(deseq2_output_gdca[[1]],
            'deseq2_outputs/', 'gdca')
```

Volcano plots! Make more efficient with a function.

```{r}
volcano <- function(deseq_results, fname, metab_name, ymax, thresh=0.10, wid=4, ht=4, label_genes=NULL, p_cap=NULL) {
  # Read in the file
  deseq_results <- read.csv(deseq_results, sep='\t')
  
  # Filter to the non-NA results
  deseq_results <- dplyr::filter(deseq_results, !is.na(padj))
  
  # Add a -log10p column
  deseq_results$neglog10p <- -log10(deseq_results$padj)
  
  # Optional capping of -log10p
  if (!is.null(p_cap)) {
    cap_val <- -log10(p_cap)            # convert p-value cap to –log10 space
    deseq_results$neglog10p <- pmin(deseq_results$neglog10p, cap_val)
    y_lab <- paste0("-Log10 adjusted p-value (capped at ", round(cap_val), ")")
  } else {
    y_lab <- "-Log10 adjusted p-value"
  }
  
  # Add a column for significance categorization
  deseq_results$significance <- ifelse(deseq_results$padj <= thresh & deseq_results$log2FoldChange >= 0.1, "Upregulated", ifelse(deseq_results$padj <= thresh & deseq_results$log2FoldChange <= -0.1, "Downregulated", "n.s."))
  
  # Define which genes to label
  deseq_results$label <- ifelse(deseq_results$Gene_ID %in% label_genes, ifelse(is.na(deseq_results$Gene_Name) | deseq_results$Gene_Name == "",
                                                                                       deseq_results$Gene_ID, deseq_results$Gene_Name),
                                NA)
  
  # Plot
  ggplot(deseq_results, aes(x = log2FoldChange, y = neglog10p, color = significance)) +
  geom_point(alpha = 0.6, size = 1.5) +  # Scatter plot points
  scale_color_manual(values = c("Upregulated" = "#9f329f", "Downregulated" = "#008888", "n.s." = "grey")) +
    geom_vline(xintercept = 0, linetype = "dashed", color = "gray70", linewidth = 0.5) +
    geom_hline(yintercept = 1, linetype = "dashed", color = "gray70", linewidth = 0.5) +
    ylim(0, ymax) +
    theme_bw() +
    theme(legend.position = "none",
        axis.text = element_text(size = 12, color = 'black'),
        axis.title = element_text(size = 14, color = 'black'),
        plot.title = element_text(size = 14, hjust = 0.5, color = 'black')) +
  labs(x = "Log2 Fold Change vs. Control",
       y = y_lab,
       title = paste0("Volcano plot for ", metab_name)) +
    geom_text_repel(aes(label = label),
                    na.rm = TRUE,
                    size = 3,
                    color = 'black',
                    segment.size = 0.4,
                    segment.color = "black",
                    force = 1.5,
                    direction = "both",
                    box.padding = 0.5,
                    point.padding = 0.3,
                    min.segment.length = 0.1,
                    max.overlaps = Inf)
  
  ggsave(fname, width=wid, height=ht)
}
```

Apply the function.

```{r}
# PA
volcano('deseq2_outputs/pa_all_results.txt', 'volcano_plot_pa.pdf',
        'PA (22.2 mg/kg)', 5,
        label_genes = c(
          "ENSMUSG00000022893.15",
          "ENSMUSG00000023034.8",
          "ENSMUSG00000026628.14",
          "ENSMUSG00000033730.5",
          "ENSMUSG00000038495.15",
          "ENSMUSG00000039826.14",
          "ENSMUSG00000120854.1",
          "ENSMUSG00000121069.1"),
wid = 4, ht = 4)

# IAA
volcano('deseq2_outputs/iaa_all_results.txt', 'volcano_plot_iaa.pdf',
        'IAA (1.5 mg/kg)', 5,
        label_genes = c(
          "ENSMUSG00000009734.19",
          "ENSMUSG00000040842.18",
          "ENSMUSG00000120777.1"),
wid = 4, ht = 4)

# GDCA
volcano('deseq2_outputs/gdca_all_results.txt', 'volcano_plot_gdca.pdf',
        'GDCA (0.4 mg/kg)', 30,
        label_genes = c("ENSMUSG00000090698.3",
                        "ENSMUSG00000028195.5",
                        "ENSMUSG00000055148.8",
                        "ENSMUSG00000022893.15",
                        "ENSMUSG00000038587.11",
                        "ENSMUSG00000028645.12",
                        "ENSMUSG00000042622.15",
                        "ENSMUSG00000046223.11",
                        "ENSMUSG00000044786.7",
                        "ENSMUSG00000020044.14",
                        "ENSMUSG00000020108.5",
                        "ENSMUSG00000026628.14",
                        "ENSMUSG00000054364.6",
                        "ENSMUSG00000038393.15",
                        "ENSMUSG00000018143.11",
                        "ENSMUSG00000003032.9",
                        "ENSMUSG00000048489.13",
                        "ENSMUSG00000032515.9",
                        "ENSMUSG00000023067.15",
                        "ENSMUSG00000020423.7"),
wid = 4, ht = 4, p_cap = 1e-30)
```