---
title: "Untitled"
output: html_document
date: "2025-08-24"
---

Load libraries.

```{r}
library(tximport)
library(DESeq2)
library(ggplot2)
library(dplyr)
library(readxl)
library(vegan)
library(pheatmap)
library(biomaRt)
```

First step is to get rlog-stabilized counts of genes per sample. A lot of the below code is setup for that.

Read in the data frame that came with the kallisto index, which links ensembl transcripts to ensembl genes.

```{r}
t2g <- read.csv('t2g.txt', sep = '\t', header = FALSE)
# Rename columns
t2g <- t2g %>%
  dplyr::rename(
    Transcript_ID = V1,
    Gene_ID = V2,
    Gene_Name = V3,
    Transcript_Name = V4,
    Chromosome = V5,
    Start_Position = V6,
    End_Position = V7,
    Strand_Direction = V8
  )

t2g_no_transcripts <- dplyr::select(t2g, Gene_ID, Gene_Name) %>% distinct()

# make a minimal version of t2g
t2g <- dplyr::select(t2g, Transcript_ID, Gene_ID)
```

tximport function that can read in the kallisto transcript-level counts for samples of interest.

```{r}
tximport_function <- function(sample_condition_df, kallisto_dir) {
  # Construct file paths for each sample’s Kallisto output
  files <- file.path(kallisto_dir, paste0('kallisto_out_', sample_condition_df$sample), "abundance.tsv")
  names(files) <- sample_condition_df$sample
  
  # Read in the corresponding kallisto results and aggregate to gene level
  txi.kallisto <- tximport(files, type = "kallisto", tx2gene = t2g, txIn = TRUE, txOut = FALSE)
  # explicitly specifying TRUE/FALSE for txIn and txOut even though they're the default according to the tximport manual
  # tx2gene needs transcript id as column 1, gene id as column 2
  
  return(txi.kallisto)
}
```

Apply it for TMAO samples!

```{r}
# Read in conditions table
conditions <- read_excel('sample_info_mouse_brain_rnaseq.xlsx')

# Make a "minimal" version, filter to TMAO samples, and filter out the outlier
conditions_minimal <- dplyr::select(conditions, Experiment,
                                    `Sample name`, Treatment)
colnames(conditions_minimal) <- c("Experiment", "sample", "condition")
tmao_conditions_minimal <- dplyr::filter(conditions_minimal, Experiment == 1) %>%
  dplyr::select(., -Experiment)
tmao_conditions_minimal <- dplyr::filter(tmao_conditions_minimal, sample != "TMAO_3")

# Apply tximport function
txi_tmao_samples <- tximport_function(tmao_conditions_minimal, '/personal/path/to/expt1/kallisto/files')
```

To get rlog stabilized counts, need to run DESeq2.

DESeq2 function below.

```{r}
deseq2_function <- function(sample_condition_df, txi, smallestGroupSize=3, a=0.10) {
  # Apparently absolutely critical for analysis that the column names
  # of the counts table from txi matches the order of samples in sample_condition_df
  # (see here - https://www.bioconductor.org/packages/release/bioc/vignettes/DESeq2/inst/doc/DESeq2.html#count-matrix-input)
  sample_condition_df <- tibble::column_to_rownames(sample_condition_df, var = 'sample')
  stopifnot(all(colnames(txi$counts) == rownames(sample_condition_df)))
  
  # Create the DESeq2 dataset
  # code from https://www.bioconductor.org/packages/release/bioc/vignettes/DESeq2/inst/doc/DESeq2.html#tximport
  dds <- DESeqDataSetFromTximport(txi, colData = sample_condition_df, design = ~condition)
  
  # Pre-filter low-count genes
  # https://www.bioconductor.org/packages/release/bioc/vignettes/DESeq2/inst/doc/DESeq2.html#pre-filtering
  keep <- rowSums(counts(dds) >= 10) >= smallestGroupSize
  dds <- dds[keep,]
  
  # Run DESeq2 analysis, return results with specified alpha level
  # https://www.bioconductor.org/packages/release/bioc/vignettes/DESeq2/inst/doc/DESeq2.html#quick-start
  dds <- DESeq(dds)
  # Extract the condition effect by name
  cond_name <- grep("^condition_", resultsNames(dds), value = TRUE)
  print(cond_name)
  dds_results <- results(dds, name = cond_name, alpha = a)
  # Shrink coefficients
  dds_results <- lfcShrink(dds, coef=cond_name, type = "apeglm", res=dds_results)
  
  # Return results and DESeq2 object
  return(list(results = dds_results, dds = dds))
}
```

Run DESeq2 and get rlog-stabilized counts.

```{r}
tmao_conditions_minimal$condition <- factor(tmao_conditions_minimal$condition, levels = c("Saline", "TMAO"))
deseq2_output <- deseq2_function(tmao_conditions_minimal, txi_tmao_samples, smallestGroupSize=3)
# https://bioconductor.org/packages/devel/bioc/vignettes/DESeq2/inst/doc/DESeq2.html#extracting-transformed-values
rlog_obj <- rlog(deseq2_output[[2]], blind = TRUE)
rlog_counts <- data.frame(assay(rlog_obj))
```

Extract rlog counts specifically for our genes of interest.

```{r}
rlog_counts <- tibble::rownames_to_column(rlog_counts, var = "ensembl_gene_id")
# Merge to get gene names
rlog_counts <- merge(rlog_counts, t2g_no_transcripts,
                     by.x = 'ensembl_gene_id',
                     by.y = 'Gene_ID')
# Define genes of interest
gaba_a_receptor_genes <- c("Gabra1", "Gabra2", "Gabra6", "Gabrb1", "Gabrb2", "Gabrb3", "Gabrg1", "Gabrg2")
gaba_a_receptor_genes_rlog <- dplyr::filter(rlog_counts, Gene_Name %in% gaba_a_receptor_genes)

# Heatmap
sample_names <- c("S_1", "S_2", "S_3", "S_4", "TMAO_1", "TMAO_2", "TMAO_4")
gaba_a_receptor_genes_rlog <- dplyr::select(gaba_a_receptor_genes_rlog, Gene_Name, all_of(sample_names))
  
# Separate gene names
gene_names <- gaba_a_receptor_genes_rlog$Gene_Name
  
# Remove gene name column for scaling
gaba_a_receptor_genes_rlog <- as.matrix(gaba_a_receptor_genes_rlog[, -1])

# Scale each row (z-score across samples)
scaled_matrix <- t(apply(gaba_a_receptor_genes_rlog, 1, scale))
  
# Restore column names
colnames(scaled_matrix) <- sample_names
  
# Add gene names back
rownames(scaled_matrix) <- gene_names
  
pheatmap(scaled_matrix,
         scale = "none",
         show_rownames = TRUE,
         show_colnames = TRUE,
         filename = 'gaba_a_receptor_genes_heatmap.pdf',
         cluster_cols = FALSE,
         height = 4)
  
```