---
title: "Untitled"
output: html_document
date: "2025-09-22"
---

Load libraries.

```{r}
library(chromVAR)
library(motifmatchr)
library(JASPAR2024)
library(BSgenome.Mmusculus.UCSC.mm10)
library(TFBSTools)
library(dplyr)
library(SummarizedExperiment)
library(ggplot2)
library(pheatmap)
```

Following guidance from: https://link.springer.com/protocol/10.1007/978-1-0716-1103-6_10
Get TF motif database.

```{r}
set.seed(123)
opts <- list()
opts["species"] <- "Mus musculus"
opts["collection"] <- "CORE"

sqlite_file <- JASPAR2024()

motifs <- TFBSTools::getMatrixSet(sqlite_file@db, opts)

# attach human readable names to motif IDs
if (!isTRUE(all.equal(TFBSTools::name(motifs), names(motifs)))) {
  names(motifs) <- paste(names(motifs), TFBSTools::name(motifs), sep = "_")
}
```

Prepare ATAC-seq data.

```{r}
genome <- BSgenome.Mmusculus.UCSC.mm10
# read in counts data frame
counts <- read.table('All_Samples.counts', sep = '\t', header = TRUE)
# Define peaks
rowRanges <- makeGRangesFromDataFrame(counts[,c(2,3,4)],
                                      seqnames.field = "Chr",
                                      start.field = "Start",
                                      end.field = "End")
# Separate numeric data
counts_only <- as.matrix(dplyr::select(counts, Brain_1_pooled.sorted.bam, Brain_2_pooled.sorted.bam, Brain_4_pooled.sorted.bam, Brain_5_pooled.sorted.bam, Brain_6_pooled.sorted.bam, Brain_7_pooled.sorted.bam, Brain_8_pooled.sorted.bam))
# Make an "experiment" containing the data
colData <- DataFrame(Condition=c("Control", "GDCA", "GDCA", "Control", "GDCA", "Control", "GDCA"), row.names=colnames(counts_only))
# Sanity check
stopifnot(all(rownames(colData) == colnames(counts_only)))
rse <- SummarizedExperiment::SummarizedExperiment(assays=SimpleList(counts=counts_only), rowRanges=rowRanges, colData=colData)
```

Enable chromVAR to correct for GC bias.

```{r}
rse_gc <- addGCBias(rse, genome = genome)
```

Match motifs to peaks.

```{r}
motif_ix <- matchMotifs(motifs, rse_gc, genome = genome)

# check structure
motif_matrix <- motifMatches(motif_ix)
motif_matrix[1:5, 1:5]  # first 5 peaks × first 5 motifs
```

Compute deviations and variability of deviations across samples.

```{r}
dev_TF <- computeDeviations(object = rse_gc, annotations = motif_ix)
variability_TF <- computeVariability(dev_TF)

# Output the resulting DF
variability_TF_output <- tibble::rownames_to_column(variability_TF,
                                                    var = 'motif')
variability_TF_output <- dplyr::rename(variability_TF_output, TF_name=name)
write.table(variability_TF_output, 'chromvar_results/TF_variability.txt', sep = '\t', quote = FALSE, row.names = FALSE)
```

Look at sample correlations.

```{r}
sample_cor_TF <- getSampleCorrelation(dev_TF)
colnames(sample_cor_TF) <- c("Control_1", "GDCA_1", "GDCA_2", "Control_2", "GDCA_3", "Control_3", "GDCA_4")
rownames(sample_cor_TF) <- colnames(sample_cor_TF)
pheatmap(sample_cor_TF, clustering_distance_rows =
     as.dist(1-sample_cor_TF),  clustering_distance_cols =
  as.dist(1-sample_cor_TF), filename = "chromvar_results/heatmap_sample_correlations.pdf",
         height=3,width=7)
```

Make a graph showing the 10 most variable TFs.

```{r}
# Read in again if necessary
variability_TF_output <- read.csv('chromvar_results/TF_variability.txt', sep = '\t')
# Order by variability
variability_TF_output <- variability_TF_output[order(variability_TF_output$variability, decreasing = TRUE),]
# Take top 10
variability_TF_top10 <- variability_TF_output[seq(1,10),]

# make motif a factor
variability_TF_top10$motif <- factor(variability_TF_top10$motif, levels = rev(variability_TF_top10$motif))

# cap padj
variability_TF_top10$p_value_adj <- pmax(variability_TF_top10$p_value_adj, 1E-200)

ggplot(variability_TF_top10, aes(x = motif, y = variability)) +
  geom_pointrange(aes(ymin = bootstrap_lower_bound, 
                      ymax = bootstrap_upper_bound, 
                      color = -log10(p_value_adj))) +
  coord_flip() +
  scale_color_gradient(low = "#99cfcf", high = "#006c6c") +
  labs(title = "Highly Variable TF Motifs",
       y = "Variability (with bootstrap CI)",
       x = "TF motif",
       color = "-log10(padj)\ncapped at 200") +
  theme_bw() +
  theme(legend.position = "right",
      axis.text = element_text(size = 8, color = 'black'),
      axis.title = element_text(size = 10, color = 'black'),
      plot.title = element_text(size = 10, color = 'black'),
      legend.text = element_text(size = 8, color = 'black'),
      legend.title = element_text(size = 8, color = 'black'))

ggsave('chromvar_results/highly_variable_tf_motifs.pdf', width = 4, height = 3)
```

Examine directionality of changes of variable TFs with heatmap - maybe take the top 25.

```{r}
# Extract per-sample deviations for all TFs
Dev_Highest_var_TF <- as.data.frame(assays(dev_TF)[[1]])
# Merge with variability_TF
Dev_Highest_var_TF <- tibble::rownames_to_column(Dev_Highest_var_TF, var = 'Motif')
variability_TF <- tibble::rownames_to_column(variability_TF, var = 'Motif')
# Merge
merged <- merge(Dev_Highest_var_TF, variability_TF, by = 'Motif')
# Subset to top 25 variable TFs
merged <- merged[order(merged$variability, decreasing = TRUE),]
merged <- merged[seq(1,25),]
# Extract numeric portion
rownames(merged) <- NULL
merged <- tibble::column_to_rownames(merged, var = 'Motif')
merged <- as.matrix(dplyr::select(merged, Brain_1_pooled.sorted.bam, Brain_2_pooled.sorted.bam, Brain_4_pooled.sorted.bam, Brain_5_pooled.sorted.bam, Brain_6_pooled.sorted.bam, Brain_7_pooled.sorted.bam, Brain_8_pooled.sorted.bam))
# Rename columns
colnames(merged) <- c("Control_1", "GDCA_1", "GDCA_2", "Control_2", "GDCA_3", "Control_3", "GDCA_4")
# Reorder columns
merged <- as.matrix(dplyr::select(data.frame(merged), "Control_1", "Control_2", "Control_3", "GDCA_1", "GDCA_2", "GDCA_3", "GDCA_4"))
pheatmap(merged, cluster_rows = F,
     cluster_cols = F, scale = "row", filename =
     "chromvar_results/Top25_TF_Heatmap.pdf",height=7,width=7)
```