---
title: "Untitled"
output: html_document
date: "2025-09-21"
---

Load libraries.

```{r setup, include=FALSE}
library(dplyr)
library(ggplot2)
library("DESeq2")
```

Read in counts data and format appropriately.

```{r}
counts <- read.table('All_Samples.counts', sep = '\t', header = TRUE)
# Get rid of extraneous columns
counts_only <- tibble::column_to_rownames(counts, var = 'Geneid')
counts_only <- dplyr::select(counts_only,
                             Brain_1_pooled.sorted.bam,
                             Brain_2_pooled.sorted.bam,
                             Brain_3_pooled.sorted.bam,
                             Brain_4_pooled.sorted.bam,
                             Brain_5_pooled.sorted.bam,
                             Brain_6_pooled.sorted.bam,
                             Brain_7_pooled.sorted.bam,
                             Brain_8_pooled.sorted.bam)
# Rename columns
counts_only <- dplyr::rename(counts_only,
                             S1=Brain_1_pooled.sorted.bam,
                             G1=Brain_2_pooled.sorted.bam,
                             S2=Brain_3_pooled.sorted.bam,
                             G2=Brain_4_pooled.sorted.bam,
                             S3=Brain_5_pooled.sorted.bam,
                             G3=Brain_6_pooled.sorted.bam,
                             S4=Brain_7_pooled.sorted.bam,
                             G4=Brain_8_pooled.sorted.bam)
# Turn into matrix
counts_only <- as.matrix(counts_only)
```

Define metadata.

```{r}
metadata <- data.frame(sample = colnames(counts_only),
                       condition = c("Saline",
                                     "GDCA",
                                     "Saline",
                                     "GDCA",
                                     "Saline",
                                     "GDCA",
                                     "Saline",
                                     "GDCA"))
metadata <- tibble::column_to_rownames(metadata, var = "sample")
metadata$condition <- factor(metadata$condition, levels =
                                  c("Saline", "GDCA"))
```

Now create the DESeq2 dataset and run DESeq2, with and without brain 3 (low TSS).

```{r}
# Including brain 3

# Sanity check
stopifnot(all(colnames(counts_only) == rownames(metadata)))
# The below line is adapted from:
# https://link.springer.com/protocol/10.1007/978-1-0716-1103-6_10
dds <- DESeqDataSetFromMatrix(countData = counts_only,
                      colData=metadata, design=~condition)
# Pre-filter low-count peaks
# https://www.bioconductor.org/packages/release/bioc/vignettes/DESeq2/inst/doc/DESeq2.html#pre-filtering
keep <- rowSums(counts(dds) >= 10) >= 4 # 4 is smallest group size
dds <- dds[keep,]
# Get results!
dds_1 <- DESeq(dds)
# Extract the condition effect by name
cond_name <- grep("^condition_", resultsNames(dds_1), value = TRUE)
dds_results <- results(dds_1, name = cond_name, alpha = 0.1)

# Merge with coordinate info
coord_info <- dplyr::select(counts, Geneid, Chr, Start, End)
dds_results <- data.frame(dds_results)
dds_results <- tibble::rownames_to_column(dds_results, var = "Geneid")
dds_results <- merge(dds_results, coord_info, by = "Geneid")
write.table(dds_results, 'deseq_results_with_brain3.tsv', sep = '\t', quote = FALSE, row.names = FALSE)

# Without brain 3 - use as final results

counts_only_no_brain_3 <- counts_only[, colnames(counts_only) != "S2"]
metadata_no_brain_3 <- data.frame(sample = colnames(counts_only_no_brain_3),
                       condition = c("Saline",
                                     "GDCA",
                                     "GDCA",
                                     "Saline",
                                     "GDCA",
                                     "Saline",
                                     "GDCA"))
metadata_no_brain_3 <- tibble::column_to_rownames(metadata_no_brain_3,
                                                  var = "sample")
metadata_no_brain_3$condition <- factor(metadata_no_brain_3$condition,
                                        levels = c("Saline", "GDCA"))
stopifnot(all(colnames(counts_only_no_brain_3) == rownames(metadata_no_brain_3)))
dds_no_brain_3 <- DESeqDataSetFromMatrix(countData = counts_only_no_brain_3,
                      colData=metadata_no_brain_3, design=~condition)
keep <- rowSums(counts(dds_no_brain_3) >= 10) >= 3 # 3 is smallest group size
dds_no_brain_3 <- dds_no_brain_3[keep,]
# Get results!
dds_no_brain_3 <- DESeq(dds_no_brain_3)
# Extract the condition effect by name
cond_name <- grep("^condition_", resultsNames(dds_no_brain_3), value = TRUE)
dds_results_no_brain_3 <- results(dds_no_brain_3, name = cond_name, alpha = 0.1)

# Merge with coordinate info
dds_results_no_brain_3 <- data.frame(dds_results_no_brain_3)
dds_results_no_brain_3 <- tibble::rownames_to_column(dds_results_no_brain_3, var = "Geneid")
dds_results_no_brain_3 <- merge(dds_results_no_brain_3, coord_info, by = "Geneid")
write.table(dds_results_no_brain_3, 'deseq_results.tsv', sep = '\t', quote = FALSE, row.names = FALSE)

# Generally not using lfcshrink because these are not RNA data
```

Get some high level stats on both sets of DESeq results.

```{r}
# Brain 3 included
sig_results <- dplyr::filter(data.frame(dds_results), padj <= 0.1)
sum(sig_results$log2FoldChange > 0)
sum(sig_results$log2FoldChange < 0)
min(abs(sig_results$log2FoldChange))

# No brain 3 (final)
sig_results_no_brain_3 <- dplyr::filter(data.frame(dds_results_no_brain_3), padj <= 0.1)
sum(sig_results_no_brain_3$log2FoldChange > 0)
sum(sig_results_no_brain_3$log2FoldChange < 0)
min(abs(sig_results_no_brain_3$log2FoldChange))

# Overlap
sig_results_minimal <- dplyr::select(sig_results, Geneid, log2FoldChange)
sig_results_minimal <- dplyr::rename(sig_results_minimal, LFC_with_brain3 = log2FoldChange)
sig_results_no_brain_3_minimal <- dplyr::select(sig_results_no_brain_3, Geneid, log2FoldChange)
sig_results_no_brain_3_minimal <- dplyr::rename(sig_results_no_brain_3_minimal, LFC_no_brain3 = log2FoldChange)
overlap <- merge(sig_results_minimal, sig_results_no_brain_3_minimal, by = "Geneid")
overlap$same_direction <- sign(overlap$LFC_with_brain3) == sign(overlap$LFC_no_brain3)

# Same direction is always true
# 190 rows in overlap
# 190 / 207 = 0.918
```

Volcano plot - results without brain 3.

```{r}
# Filter to the non-NA results
dds_results_no_brain_3 <- dplyr::filter(dds_results_no_brain_3, !is.na(padj))

# Add a -log10p column
dds_results_no_brain_3$neglog10p <- -log10(dds_results_no_brain_3$padj)

# Add a column for significance categorization
dds_results_no_brain_3$significance <- ifelse(dds_results_no_brain_3$padj <= 0.1 & dds_results_no_brain_3$log2FoldChange > 0, "More accessible", ifelse(dds_results_no_brain_3$padj <= 0.1 & dds_results_no_brain_3$log2FoldChange < 0, "Less accessible", "Not significant"))

ggplot(dds_results_no_brain_3, aes(x = log2FoldChange, y = neglog10p, color = significance)) +
geom_point(alpha = 0.6, size = 1.5) +
scale_color_manual(values = c("More accessible" = "#9f329f", "Less accessible" = "#008888", "Not significant" = "gray")) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray70", linewidth = 0.5) +
    geom_hline(yintercept = 1, linetype = "dashed", color = "gray70", linewidth = 0.5) +
theme_bw() +
labs(title = "Volcano plot for GDCA ATAC-Seq", x = "Log2 Fold Change vs. Control", y = "-Log10 Adjusted p-value", color = "Significance") +
theme(legend.position = "none",
      axis.text = element_text(size = 8, color = 'black'),
      axis.title = element_text(size = 10, color = 'black'),
      plot.title = element_text(size = 10, color = 'black')) 

ggsave('volcano_plot.pdf', width = 3.75, height = 2.75)
```

Output files that are needed for the coverage heatmaps (section 3.9 from this article: https://link.springer.com/protocol/10.1007/978-1-0716-1103-6_10)

```{r}
# Filter the significant results (without brain 3) to "up" and "down" peaks, and select only important columns
sig_results_no_brain_3_up <- dplyr::filter(sig_results_no_brain_3, log2FoldChange > 0) %>% dplyr::select(., Chr, Start, End, baseMean, log2FoldChange)
sig_results_no_brain_3_down <- dplyr::filter(sig_results_no_brain_3, log2FoldChange < 0) %>% dplyr::select(., Chr, Start, End, baseMean, log2FoldChange)

# Convert to bed format
sig_results_no_brain_3_up$Start <- sig_results_no_brain_3_up$Start-1
sig_results_no_brain_3_down$Start <- sig_results_no_brain_3_down$Start-1

# Write out the tables
write.table(sig_results_no_brain_3_up, 'up_peaks_input_for_cov_heatmaps.bed', sep = '\t', quote = FALSE, row.names = FALSE)

write.table(sig_results_no_brain_3_down, 'down_peaks_input_for_cov_heatmaps.bed', sep = '\t', quote = FALSE, row.names = FALSE)
```

For HOMER peak annotation of differential peaks.

```{r}
# Modify sig results to just the needed columns
sig_results_no_brain_3 <- dplyr::select(sig_results_no_brain_3, Geneid, Chr, Start, End)
# add "strand" information
sig_results_no_brain_3$Strand <- "."
# write file
write.table(sig_results_no_brain_3, "sig_peak_file_for_homer.txt", sep='\t', quote = FALSE, row.names = FALSE, col.names = FALSE)
```