---
title: "Untitled"
output: html_document
date: "2025-09-21"
---

Load libraries.

```{r}
library(dplyr)
library(ggplot2)
library(ggrepel)
library(see)
library(rGREAT)
library(GenomicRanges)
library(rtracklayer)
library(GenomeInfoDb)
library(TxDb.Mmusculus.UCSC.mm10.knownGene)
library(tidyr)
```

First, examine the annotation stats.

```{r}
ann_stats <- read.csv('annotation_stats.txt', sep = '\t')

# I really care about the basic annotation stats
ann_stats <- ann_stats[seq(1,12),]
# Sanity check
stopifnot(sum(as.numeric(ann_stats$Number.of.peaks)) == 207)
```

First, make a pie chart showing the % of differential peaks in each category.

```{r}
# Filter to rows with number of peaks > 0
ann_stats$Number.of.peaks <- as.numeric(ann_stats$Number.of.peaks)
ann_stats_present_peaks <- dplyr::filter(ann_stats, Number.of.peaks > 0)

# Compute percentages and label position
ann_stats_present_peaks <- ann_stats_present_peaks %>%
  mutate(
    Percentage = round(100 * Number.of.peaks / sum(Number.of.peaks), 1),
    Fraction = Number.of.peaks / sum(Number.of.peaks),
    Label = paste0(Annotation, " (", Percentage, "%)"),
    ypos = cumsum(Fraction) - 0.5 * Fraction  # midpoint of each slice
  )

# Color blind friendly palette
cb_palette <- see::okabeito_colors()[seq(1,6)]
names(cb_palette) <- c("Intergenic", "Promoter", "Intron", "3UTR", "TTS", "Exon")

# Plot with labels outside and connecting lines
ggplot(ann_stats_present_peaks, aes(x = 1, y = Fraction, fill = Annotation)) +
  geom_col(width = 1, color = "white") +
  coord_polar("y") +
  theme_void() +
  labs(title = "Differential Peaks: Overlapping Genomic Features") +
  scale_fill_manual(values = cb_palette) +
  theme(legend.position = "none") +
  geom_text_repel(
    aes(y = ypos, label = Label),
    nudge_x = 1.8,
    segment.size = 0.5,
    show.legend = FALSE,
    size = 4,
    direction = "y",
    hjust = 0
  ) +
  xlim(0.5, 2.8)  # space for labels outside

ggsave('pc_peaks_per_major_feature.pdf', width = 7, height = 4)
```

Next make a bar chart showing the log2 ratio (obs/expected) peaks.
Annotate with significance stars.

```{r}
ann_stats_present_peaks <- dplyr::select(ann_stats_present_peaks,
                                         Number.of.peaks,
                                         Annotation,
                                         Log2.Ratio..obs.exp.,
                                         LogP.enrichment...values.depleted.)

# Rename columns above
ann_stats_present_peaks <- dplyr::rename(ann_stats_present_peaks,
                                         Obs_Exp_Ratio=Log2.Ratio..obs.exp.,
                                         Signed_logP=LogP.enrichment...values.depleted.)

# Make relevant columns numeric
ann_stats_present_peaks$Obs_Exp_Ratio <- as.numeric(ann_stats_present_peaks$Obs_Exp_Ratio)

ann_stats_present_peaks$Signed_logP <- as.numeric(ann_stats_present_peaks$Signed_logP)

# Add significance, significance stars, and directionality
# I know the log for the p-value was a natural log based on trying the binomial test myself for a couple rows
ann_stats_present_peaks <- ann_stats_present_peaks %>%
  dplyr::mutate(
    pval = exp(-abs(Signed_logP)), # log of p-value between 0 and 1 will always be negative so have to make sure you're including that negative in the exponentiation
    direction = ifelse(Signed_logP < 0, "Obs > Exp", "Obs < Exp"), # because + values are depleted according to the original column header
    stars = case_when(
      pval < 0.001 ~ "***",
      pval < 0.01  ~ "**",
      pval < 0.05  ~ "*",
      TRUE ~ ""
    )
  )

# Reorder
ann_stats_present_peaks <- ann_stats_present_peaks[order(ann_stats_present_peaks$Obs_Exp_Ratio),]

# Make relevant columns into factors
ann_stats_present_peaks$Annotation <- factor(ann_stats_present_peaks$Annotation, levels = ann_stats_present_peaks$Annotation)
ann_stats_present_peaks$direction <- factor(ann_stats_present_peaks$direction, levels = c("Obs < Exp", "Obs > Exp"))

# Assign positions to the significance stars
ann_stats_present_peaks <- dplyr::mutate(ann_stats_present_peaks,
                                         star_y = ifelse(Obs_Exp_Ratio > 0,
                                                         Obs_Exp_Ratio + 0.05,
                                                         Obs_Exp_Ratio - 0.05), star_vjust = ifelse(Obs_Exp_Ratio > 0, 0, 1))

# Now plot
ggplot(ann_stats_present_peaks, aes(x = Annotation,
               y = Obs_Exp_Ratio,
               fill = direction)) + geom_col(width = 0.7) + 
  geom_hline(yintercept = 0, linewidth = 0.6, colour = "black") +
  geom_text(aes(y = star_y, label = stars, vjust = star_vjust),
            size = 5, fontface = "bold") +
  scale_fill_manual(values = c(`Obs < Exp` = "#008888",
                               `Obs > Exp` = "#9f329f")) +
  labs(x = NULL, y = "Log2(Observed / Expected)", fill = "Direction",
       title = "Differential Peaks: Overlapping Genomic Features") +
  theme_bw() +
  theme(panel.grid = element_blank(),
        axis.text.x = element_text(angle = 45, hjust = 1, color = 'black', size = 10),
        axis.text.y = element_text(size = 8, color = 'black'),
        axis.title = element_text(size = 10, color = 'black'),
        legend.title = element_blank(),
        legend.text = element_text(size = 10, color = 'black'),
          legend.position = c(0.02, 0.98),
  legend.justification = c("left", "top"),
  plot.title = element_text(size = 10, color = 'black')) + ylim(-1,1.75)

ggsave('obs_vs_expected_per_major_feature.pdf', width = 3.75, height = 3)
```
Determine the genes whose regulatory regions overlap with each peak, using rGREAT.

How many cases do we see where:
- a region is differentially accessible between GDCA and saline brains by ATAC
- a gene associated with that region is differentially expressed by RNA-seq
- directionality of change is the same

Relevant pages describing rGREAT:
https://jokergoo.github.io/rGREAT/articles/local-GREAT.html
https://jokergoo.github.io/rGREAT/articles/other-organisms.html

```{r}
# Read in the differential accessibility results
atac_deseq_results <- read.csv('deseq_results.tsv', sep = '\t')

# Define differentially accessible peaks
diff_access <- dplyr::filter(atac_deseq_results, padj <= 0.10)
diff_access <- dplyr::select(diff_access, Geneid, Chr, Start, End)

# Define a set of background peaks
bg <- dplyr::select(atac_deseq_results, Geneid, Chr, Start, End)

# Genomic range coordinates (the necessary input to rGREAT) are 1-based inclusive so we're good on coordinates
# Sources below
# https://www.bioconductor.org/help/course-materials/2014/SeattleFeb2014/Ranges.pdf -- 1-based, closed (inclusive)
# https://bioconductor.org/packages//2.11/bioc/vignettes/rtracklayer/inst/doc/rtracklayer.pdf -- 1-based

# Convert to GRanges
# https://www.bioconductor.org/packages/devel/bioc/vignettes/GenomicRanges/inst/doc/GenomicRangesIntroduction.html
bg_gr <- GRanges(
    seqnames = bg$Chr,
    ranges = IRanges(
        start = bg$Start,
        end = bg$End,
        names = bg$Geneid
    )
)

diff_access_gr <- GRanges(
    seqnames = diff_access$Chr,
    ranges = IRanges(
        start = diff_access$Start,
        end = diff_access$End,
        names = diff_access$Geneid
    )
)

genome(bg_gr) <- "mm10"
genome(diff_access_gr) <- "mm10"

# Run GREAT
res_bp <- great(
    diff_access_gr,
    gene_sets = "GO:BP",
    tss_source = "TxDb.Mmusculus.UCSC.mm10.knownGene",
    background = bg_gr
)

enrich_tab <- getEnrichmentTable(res_bp)
peak_gene <- getRegionGeneAssociations(res_bp)

# Get the genes per peak
genes_per_peak_long <- data.frame(
  peak_name = rep(names(peak_gene), lengths(peak_gene$annotated_genes)),
  seqnames = rep(as.character(seqnames(peak_gene)), lengths(peak_gene$annotated_genes)),
  start = rep(start(peak_gene), lengths(peak_gene$annotated_genes)),
  end = rep(end(peak_gene), lengths(peak_gene$annotated_genes)),
  gene_raw = unlist(as.list(peak_gene$annotated_genes), use.names = FALSE),
  dist_to_TSS = unlist(as.list(peak_gene$dist_to_TSS), use.names = FALSE)
)

# Add LFC and padj from DESeq2 (ATAC)
atac_deseq_results_minimal <- dplyr::select(atac_deseq_results,
                                            Geneid, log2FoldChange, padj)
genes_per_peak_long <- merge(genes_per_peak_long,
                             atac_deseq_results_minimal, by.x = "peak_name",
                             by.y = "Geneid")
genes_per_peak_long <- dplyr::rename(genes_per_peak_long,
                                     log2FoldChange_ATAC=log2FoldChange,
                                     padj_ATAC=padj)
# also add an absolute distance to TSS
genes_per_peak_long$abs_dist_to_TSS <- abs(genes_per_peak_long$dist_to_TSS)

# Add LFC and padj from DESeq2 (RNA)
rna_deseq_results <- read.csv('../Mouse brain RNAseq/deseq2_outputs/gdca_all_results.txt', sep = '\t')
rna_deseq_results_minimal <- dplyr::select(rna_deseq_results,
                                           Gene_Name, log2FoldChange, padj)
rna_deseq_results_minimal <- dplyr::rename(rna_deseq_results_minimal,
                                           log2FoldChange_RNA=log2FoldChange,
                                     padj_RNA=padj)
genes_per_peak_long <- genes_per_peak_long %>%
  left_join(rna_deseq_results_minimal, by = c("gene_raw" = "Gene_Name"))

genes_per_peak_long$both_rna_and_atac_sig <- ifelse(genes_per_peak_long$padj_RNA <= 0.10, "yes", "no")

genes_per_peak_long$directionality_rna_vs_atac <- ifelse(genes_per_peak_long$both_rna_and_atac_sig == "yes", ifelse(sign(genes_per_peak_long$log2FoldChange_ATAC) == sign(genes_per_peak_long$log2FoldChange_RNA), "same", "different"), NA)
```

Add in some of the information from HOMER, then output.

```{r}
homer_annotations <- read.csv('annotated_sig_peak_file.txt', sep = '\t')
colnames(homer_annotations)[1] <- "Peak_id"

homer_annotations <- dplyr::select(homer_annotations,
                                   Peak_id, Chr, Start, End, Annotation,
                                   Detailed.Annotation)
homer_annotations <- dplyr::rename(homer_annotations,
                                   HOMER_annotation = Annotation,
                                   HOMER_detailed_annotation = Detailed.Annotation)

genes_per_peak_long <- dplyr::rename(genes_per_peak_long,
                                     Peak_id = peak_name,
                                     Chr = seqnames,
                                     Start = start,
                                     End = end,
                                     Overlaps_reg_region_of_this_gene = gene_raw)

merged <- merge(genes_per_peak_long, homer_annotations,
                by = c("Peak_id", "Chr", "Start", "End"))

# Reorder columns
merged <- dplyr::select(merged, Peak_id, Chr, Start, End, log2FoldChange_ATAC,
                        padj_ATAC, HOMER_annotation,
                        HOMER_detailed_annotation, Overlaps_reg_region_of_this_gene,
                        dist_to_TSS, abs_dist_to_TSS,
                        log2FoldChange_RNA, padj_RNA, both_rna_and_atac_sig,
                        directionality_rna_vs_atac)

merged <- dplyr::rename(merged,
                        Overlaps_reg_region_of_this_gene_GREAT=Overlaps_reg_region_of_this_gene,
                        dist_to_TSS_GREAT=dist_to_TSS,
                        abs_dist_to_TSS_GREAT=abs_dist_to_TSS)

write.table(merged, 'annotated_sig_peaks_file_with_homer_and_great_info.txt', sep = '\t', quote = FALSE, row.names = FALSE)
```