---
title: "Untitled"
output: html_document
date: "2025-08-14"
---

Load necessary libraries.

```{r}
library("dplyr")
library("tximport")
library(ggplot2)
library("DESeq2")
library("rhdf5")
library("apeglm")
library("readxl")
library(stringr)
library(corrplot)
library(Biostrings)
```

First, generate a data frame that links ensembl transcripts to ensembl genes - will need for tximport.

Originally, I was using biomart. However, as of 2026, Ensembl is suffering from server issues.

So, using the fasta file from which we built the kallisto index, might be safest.

Downloaded it here (Homo_sapiens.GRCh38.cdna.all.fa.gz)

```{r}
fa <- readDNAStringSet("Homo_sapiens.GRCh38.cdna.all.fa.gz")

headers <- names(fa)

t2g <- tibble(
  TXNAME = str_extract(headers, "^\\S+"),
  GENEID = str_extract(headers, "gene:ENSG[0-9]+\\.[0-9]+"),
  SYMBOL = str_extract(headers, "gene_symbol:[^ ]+")
) %>%
  mutate(
    GENEID = sub("^gene:", "", GENEID),
    SYMBOL = sub("^gene_symbol:", "", SYMBOL)
  ) %>%
  filter(!is.na(TXNAME), !is.na(GENEID)) %>%
  distinct()

# needed later when outputting DESeq2 results
t2g_no_transcripts <- dplyr::select(t2g, GENEID, SYMBOL) %>% distinct()

t2g <- dplyr::select(t2g, TXNAME, GENEID)
```

Read in the kallisto-generated estimated counts using tximport.

https://bioconductor.org/packages/devel/bioc/vignettes/tximport/inst/doc/tximport.html#kallisto_with_TSV_files

Write a tximport function that can read in the kallisto transcript-level counts for samples of interest.

```{r}
tximport_function <- function(sample_condition_df, kallisto_dir) {
  # Construct file paths for each sample’s Kallisto output
  files <- file.path(kallisto_dir, paste0('kallisto_out_', sample_condition_df$sample), "abundance.tsv")
  names(files) <- sample_condition_df$sample
  
  # Read in the corresponding kallisto results and aggregate to gene level
  txi.kallisto <- tximport(files, type = "kallisto", tx2gene = t2g, txIn = TRUE, txOut = FALSE)
  # explicitly specifying TRUE/FALSE for txIn and txOut even though they're the default according to the tximport manual
  # tx2gene needs transcript id as column 1, gene id as column 2
  # we are ignoring Tx version because the t2g table doesn't have Tx version
  
  return(txi.kallisto)
}
```

Write a DESeq2 function.

```{r}
deseq2_function <- function(sample_condition_df, conditions_of_interest, kallisto_dir, smallestGroupSize=3, a=0.10) {
  # Filter sample_condition_df to conditions of interest
  sample_condition_df <- dplyr::filter(sample_condition_df, condition %in% conditions_of_interest)
  sample_condition_df$condition <- droplevels(sample_condition_df$condition)
  
  # Critical check to ensure correct directionality
  stopifnot(levels(sample_condition_df$condition)[1] %in% c("DMSO control", "H2O control"))
  stopifnot(length(levels(sample_condition_df$condition)) == 2)
  
  # Use the tximport function to enable getting the gene-level counts
  txi <- tximport_function(sample_condition_df, kallisto_dir)
  
  # Apparently absolutely critical for analysis that the column names
  # of the counts table from txi matches the order of samples in sample_condition_df
  # (see here - https://www.bioconductor.org/packages/release/bioc/vignettes/DESeq2/inst/doc/DESeq2.html#count-matrix-input)
  sample_condition_df <- tibble::column_to_rownames(sample_condition_df, var = 'sample')
  stopifnot(all(colnames(txi$counts) == rownames(sample_condition_df)))
  
  # Create the DESeq2 dataset
  # code from https://www.bioconductor.org/packages/release/bioc/vignettes/DESeq2/inst/doc/DESeq2.html#tximport
  dds <- DESeqDataSetFromTximport(txi, colData = sample_condition_df, design = ~condition)
  
  # Pre-filter low-count genes
  # https://www.bioconductor.org/packages/release/bioc/vignettes/DESeq2/inst/doc/DESeq2.html#pre-filtering
  keep <- rowSums(counts(dds) >= 10) >= smallestGroupSize
  dds <- dds[keep,]
  
  # Run DESeq2 analysis, return results with specified alpha level
  # https://www.bioconductor.org/packages/release/bioc/vignettes/DESeq2/inst/doc/DESeq2.html#quick-start
  dds <- DESeq(dds)
  
  coef_name <- resultsNames(dds)[2]
  message("Using coefficient: ", coef_name)
  
  stopifnot(
  length(coef_name) == 1,
  startsWith(coef_name, "condition"))
  
  dds_results <- results(dds, name = coef_name, alpha = a)
  
  dds_results <- lfcShrink(
  dds,
  coef = coef_name,
  type = "apeglm",
  res = dds_results)
  
  # Return results and DESeq2 object
  return(list(results = dds_results, dds = dds))
}
```

Read in all the conditions dfs, filter out outliers we previously identified, group relevant groups of controls, set factor levels so that DESeq2 will make the proper comparisons.

```{r}
# Batch 1
batch1_conditions <- read_excel("sample_info_transcriptomic_screen.xlsx") %>%
  dplyr::filter(., Batch == 1)
batch1_conditions <- dplyr::select(batch1_conditions, `Sample name`, Treatment)
batch1_conditions <- dplyr::rename(batch1_conditions, sample = `Sample name`, condition = Treatment)

batch1_conditions$condition <- factor(batch1_conditions$condition, levels = unique(batch1_conditions$condition))

# Batch 2
batch2_conditions <- read_excel("sample_info_transcriptomic_screen.xlsx") %>%
  dplyr::filter(., Batch == 2)
batch2_conditions <- dplyr::select(batch2_conditions, `Sample name`, Treatment)
batch2_conditions <- dplyr::rename(batch2_conditions, sample = `Sample name`, condition = Treatment)

batch2_conditions$condition <- factor(batch2_conditions$condition, levels = unique(batch2_conditions$condition))

# filter out outliers previously identified during PCA/QC analysis
batch2_conditions <- dplyr::filter(batch2_conditions, !(sample %in% c('Oxalate_1', 'Carnitine_2')))

# Batch 3
batch3_conditions <- read_excel("sample_info_transcriptomic_screen.xlsx") %>%
  dplyr::filter(., Batch == 3)
batch3_conditions <- dplyr::select(batch3_conditions, `Sample name`, Treatment)
batch3_conditions <- dplyr::rename(batch3_conditions, sample = `Sample name`, condition = Treatment)

batch3_conditions$condition <- factor(batch3_conditions$condition, levels = unique(batch3_conditions$condition))

# filter out outliers previously identified during PCA/QC analysis
batch3_conditions <- dplyr::filter(batch3_conditions, !(sample %in% c('GDCA_1', 'Quin_2')))

# Batch 4
batch4_conditions <- read_excel("sample_info_transcriptomic_screen.xlsx") %>%
  dplyr::filter(., Batch == 4)
batch4_conditions <- dplyr::select(batch4_conditions, `Sample name`, Treatment)
batch4_conditions <- dplyr::rename(batch4_conditions, sample = `Sample name`, condition = Treatment)

batch4_conditions$condition <- factor(batch4_conditions$condition, levels = unique(batch4_conditions$condition))

# filter out outliers previously identified during PCA/QC analysis
batch4_conditions <- dplyr::filter(batch4_conditions, !(sample == 'DMSO_5'))
```

Function to output DEGs.

Output multiple versions.
1) no changes, just filter to the significant results (padj <= 0.1 and abs(LFC) >= 0.1)
2) remove lines that are duplicates of each other besides the ENSG ID - basically this occurs when two different ENSG ID are mapped to the same gene
3) "" + filter to LFC > 0 (upregulated genes)
4) "" + filter to LFC < 0 (downregulated genes)

Also output a file with *all* the results, not just the significant ones.

```{r}
output_degs <- function(dds_results, outdir, outfprefix, alpha=0.10) {
  dds_results <- tibble::rownames_to_column(data.frame(dds_results), var = 'GENEID')
  
  dds_results <- merge(dds_results, t2g_no_transcripts, by = 'GENEID', all.x=TRUE, all.y=FALSE) # this adds the external gene names
  
  # Ensure the relevant columns are numeric
  dds_results$padj <- as.numeric(dds_results$padj)
  dds_results$log2FoldChange <- as.numeric(dds_results$log2FoldChange)
  
  # filter to the significant results
  sig <- dplyr::filter(dds_results, padj <= alpha, abs(log2FoldChange) >= 0.1)
  
  # Version 1
  outfname_1 <- paste0(outdir, outfprefix, '_sig.txt')
  write.table(sig, outfname_1, sep='\t', row.names=FALSE, quote=FALSE)
  
  # Version 2
  outfname_2 <- paste0(outdir, outfprefix, '_sig_no_duplicates.txt')
  dup_rows <- duplicated(sig[,-1])
  cleaned_sig <- sig[!dup_rows,]
  write.table(cleaned_sig, outfname_2, sep='\t', row.names=FALSE, quote=FALSE)
  
  # Version 3
  outfname_3 <- paste0(outdir, outfprefix, '_sig_no_duplicates_UP.txt')
  write.table(dplyr::filter(cleaned_sig, log2FoldChange > 0),
              outfname_3, sep='\t', row.names=FALSE, quote=FALSE)
  
  # Version 4
  outfname_4 <- paste0(outdir, outfprefix, '_sig_no_duplicates_DOWN.txt')
  write.table(dplyr::filter(cleaned_sig, log2FoldChange < 0),
              outfname_4, sep='\t', row.names=FALSE, quote=FALSE)
  
  # Version 5
  outfname_5 <- paste0(outdir, outfprefix, '_all_results.txt')
  write.table(dds_results, outfname_5, sep='\t', row.names=FALSE, quote=FALSE)
}
```

Run the DESeq2 function, and the function to output DEGs!

For each batch, will have a wrapper that will make our analysis easier.

Batch 1 wrapper.

```{r}
deseq2_wrapper_batch1 <- function(conditions_of_interest, fprefix, min_group_size=3) {
  deseq2_output <- deseq2_function(batch1_conditions, conditions_of_interest, '/personal/path/to/batch1/kallisto/files', smallestGroupSize=min_group_size)
  output_degs(deseq2_output[["results"]], 'deseq2_outputs/batch1/', fprefix)
  return(deseq2_output)
}
```

Batch 2 wrapper.

```{r}
deseq2_wrapper_batch2 <- function(conditions_of_interest, fprefix, min_group_size=3) {
  deseq2_output <- deseq2_function(batch2_conditions, conditions_of_interest, '/personal/path/to/batch2/kallisto/files', smallestGroupSize=min_group_size)
  output_degs(deseq2_output[["results"]], 'deseq2_outputs/batch2/', fprefix)
  return(deseq2_output)
}
```

Batch 3 wrapper.

```{r}
deseq2_wrapper_batch3 <- function(conditions_of_interest, fprefix, min_group_size=3) {
  deseq2_output <- deseq2_function(batch3_conditions, conditions_of_interest, '/personal/path/to/batch3/kallisto/files', smallestGroupSize=min_group_size)
  output_degs(deseq2_output[["results"]], 'deseq2_outputs/batch3/', fprefix)
  return(deseq2_output)
}
```

Batch 4 wrapper.

```{r}
deseq2_wrapper_batch4 <- function(conditions_of_interest, fprefix, min_group_size=3) {
  deseq2_output <- deseq2_function(batch4_conditions, conditions_of_interest, '/personal/path/to/batch4/kallisto/files', smallestGroupSize=min_group_size)
  output_degs(deseq2_output[["results"]], 'deseq2_outputs/batch4/', fprefix)
  return(deseq2_output)
}
```

Apply the wrappers!

Batch 1.

```{r}
# PAGln
pagln <- deseq2_wrapper_batch1(c('DMSO control', 'Phenylacetylglutamine'), 'pagln')

# Cholate
cholate <- deseq2_wrapper_batch1(c('DMSO control', 'Cholate'), 'cholate')

# Indoleacetate
indoleacetate <- deseq2_wrapper_batch1(c('DMSO control', 'Indoleacetate'), 'indoleacetate')

# ImP
imp <- deseq2_wrapper_batch1(c('H2O control', 'Imidazole propionate'), 'imp')

# LPS
lps_1 <- deseq2_wrapper_batch1(c('H2O control', 'LPS control'), 'lps')

# pCS
pcs <- deseq2_wrapper_batch1(c('H2O control', 'p-Cresol sulfate'), 'pcs')

# Hippurate
hippurate <- deseq2_wrapper_batch1(c('H2O control', 'Hippurate'), 'hippurate')
```

Batch 2.

```{r}
# Oxalate
oxalate <- deseq2_wrapper_batch2(c('H2O control', 'Oxalate'), 'oxalate', min_group_size=2)

# Glycerate
glycerate <- deseq2_wrapper_batch2(c('H2O control', 'Glycerate'), 'glycerate')

# Isoleucine
isoleucine <- deseq2_wrapper_batch2(c('H2O control', 'Isoleucine'), 'isoleucine')

# Valine
valine <- deseq2_wrapper_batch2(c('H2O control', 'Valine'), 'valine')

# Glycine
glycine <- deseq2_wrapper_batch2(c('H2O control', 'Glycine'), 'glycine')

# Carnitine
carnitine <- deseq2_wrapper_batch2(c('H2O control', 'Carnitine'), 'carnitine', min_group_size=2)

# LPS
lps_2 <- deseq2_wrapper_batch2(c('H2O control', 'LPS control'), 'lps')

# BHBA
bhba <- deseq2_wrapper_batch2(c('H2O control', '3-hydroxybutyrate'), 'bhba')

# Alpha-ketoglutarate
alpha_ketoglutarate <- deseq2_wrapper_batch2(c('H2O control', 'Alpha-ketoglutarate'), 'alpha_ketoglutarate')
```

Batch 3.

```{r}
# Glycodeoxycholate
glycodeoxycholate <- deseq2_wrapper_batch3(c('H2O control', 'Glycodeoxycholate'), 'glycodeoxycholate', min_group_size=2)

# Nicotinamide
nicotinamide <- deseq2_wrapper_batch3(c('H2O control', 'Nicotinamide'), 'nicotinamide')

# Pantothenate
pantothenate <- deseq2_wrapper_batch3(c('H2O control', 'Pantothenate'), 'pantothenate')

# Quinolinate
quinolinate <- deseq2_wrapper_batch3(c('H2O control', 'Quinolinate'), 'quinolinate', min_group_size=2)

# Malate
malate <- deseq2_wrapper_batch3(c('H2O control', 'Malate'), 'malate')

# TMAO
tmao <- deseq2_wrapper_batch3(c('H2O control', 'Trimethylamine N-oxide'), 'tmao')

# LPS
lps_3 <- deseq2_wrapper_batch3(c('H2O control', 'LPS control'), 'lps')

# Choline
choline <- deseq2_wrapper_batch3(c('H2O control', 'Choline'), 'choline')

# Citrulline
citrulline <- deseq2_wrapper_batch3(c('H2O control', 'Citrulline'), 'citrulline')
```

Batch 4.

```{r}
# Allantoin
allantoin <- deseq2_wrapper_batch4(c('DMSO control', 'Allantoin'), 'allantoin')

# Methyl-indoleacetate
methyl_indoleacetate <- deseq2_wrapper_batch4(c('DMSO control', 'Methyl indole-3-acetate'), 'methyl_indoleacetate')

# Palmitate
palmitate <- deseq2_wrapper_batch4(c('DMSO control', 'Palmitate'), 'palmitate')

# Indolelactate
indolelactate <- deseq2_wrapper_batch4(c('DMSO control', 'Indolelactate'), 'indolelactate')

# N-acetylalanine
nacetylalanine <- deseq2_wrapper_batch4(c('DMSO control', 'N-acetylalanine'), 'nacetylalanine')

# Hypoxanthine
hypoxanthine <- deseq2_wrapper_batch4(c('DMSO control', 'Hypoxanthine'), 'hypoxanthine')

# LPS
lps_4 <- deseq2_wrapper_batch4(c('H2O control', 'LPS control'), 'lps')

# Urate
urate <- deseq2_wrapper_batch4(c('H2O control', 'Urate'), 'urate')

# Lysine
lysine <- deseq2_wrapper_batch4(c('H2O control', 'Lysine'), 'lysine')
```

Plot number of DEGs per metabolite (removing duplicates).

```{r}
# Initialize
first_table <- read.csv('deseq2_outputs/batch1/pagln_sig_no_duplicates.txt', sep='\t')
num_degs_per_metab_no_dups <- data.frame(metab=c('pagln'), num_degs=nrow(first_table))

batch1_metab <- c('cholate', 'indoleacetate', 'imp', 'pcs', 'hippurate')
# removed pagln because already accounted for it

batch2_metab <- c('oxalate', 'glycerate', 'isoleucine', 'valine',
            'glycine', 'carnitine', 'bhba', 'alpha_ketoglutarate')

batch3_metab <- c('glycodeoxycholate', 'nicotinamide', 'pantothenate',
            'quinolinate', 'malate', 'tmao', 'choline', 'citrulline')

batch4_metab <- c('allantoin', 'methyl_indoleacetate', 'palmitate',
            'indolelactate', 'nacetylalanine', 'hypoxanthine',
            'urate', 'lysine')

add_to_num_degs_per_metab_no_dups <- function(batch, metab_name, df) {
  table_to_read <- read.csv(paste0('deseq2_outputs/', batch, '/', metab_name, '_sig_no_duplicates.txt'), sep='\t')
  df <- rbind(df, c(metab=metab_name, num_degs=nrow(table_to_read)))
  return(df)
}

for (metabolite in batch1_metab) {
  num_degs_per_metab_no_dups <- add_to_num_degs_per_metab_no_dups('batch1', metabolite, num_degs_per_metab_no_dups)
}

for (metabolite in batch2_metab) {
  num_degs_per_metab_no_dups <- add_to_num_degs_per_metab_no_dups('batch2', metabolite, num_degs_per_metab_no_dups)
}

for (metabolite in batch3_metab) {
  num_degs_per_metab_no_dups <- add_to_num_degs_per_metab_no_dups('batch3', metabolite, num_degs_per_metab_no_dups)
}

for (metabolite in batch4_metab) {
  num_degs_per_metab_no_dups <- add_to_num_degs_per_metab_no_dups('batch4', metabolite, num_degs_per_metab_no_dups)
}

stopifnot(length(unique(num_degs_per_metab_no_dups$metab)) == 30)

num_degs_per_metab_no_dups$num_degs <- as.numeric(num_degs_per_metab_no_dups$num_degs)
```

Plot.

```{r}
metabolites <- c('Phenylacetylglutamine', 'Cholate', 'Indoleacetate', 'Imidazole propionate', 'p-Cresol sulfate', 'Hippurate', 'Oxalate*', 'Glycerate', 'Isoleucine', 'Valine', 'Glycine', 'Carnitine*', '3-hydroxybutyrate', 'Alpha-ketoglutarate', 'Glycodeoxycholate*', 'Nicotinamide', 'Pantothenate', 'Quinolinate*', 'Malate', 'Trimethylamine N-oxide', 'Choline', 'Citrulline', 'Allantoin', 'Methyl indole-3-acetate', 'Palmitate', 'Indolelactate', 'N-acetylalanine', 'Hypoxanthine', 'Urate', 'Lysine')

num_degs_per_metab_no_dups$nice_metab <- metabolites

# Sort num_degs_per_metab_no_dups in reverse order
num_degs_per_metab_no_dups <- num_degs_per_metab_no_dups[order(num_degs_per_metab_no_dups$num_degs, decreasing=TRUE),]
# make nice_metab into factor
num_degs_per_metab_no_dups$nice_metab <- factor(num_degs_per_metab_no_dups$nice_metab, levels = num_degs_per_metab_no_dups$nice_metab)

ggplot(num_degs_per_metab_no_dups, aes(x=nice_metab, y=num_degs)) + geom_bar(stat='identity', fill='#66b7b7') + labs(y='# DEGs (DESeq2)') + theme_bw() + 
    theme(axis.line = element_line(colour = "black"), 
        axis.title.y=element_text(size=18, color='black'),
        axis.title.x=element_blank(),
        axis.text.x=element_text(size=16, color = "black", angle=90, hjust=1, vjust=0.5), 
        axis.text.y=element_text(size=16, color='black'),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank()) + scale_y_log10() + geom_text(aes(label = num_degs), vjust = 0.5, hjust=1, angle=90, size = 6, color = 'black') 

ggsave('num_degs_per_metabolite.pdf', height=6.5, width=11)
```

Corrplot - sanity check that same-solvent controls on different plates can be grouped.

```{r}
# Batch 1 - H2O controls on different plates
batch_1_h2o <- tximport_function(dplyr::filter(batch1_conditions, condition == 'H2O control'), '/personal/path/to/batch1/kallisto/files')
# Log transform the gene abundances
abundances <- log(batch_1_h2o$abundance + 0.5) # using same log transformation we used for PCA
correlation_matrix <- cor(abundances)
pdf('corrplot/batch1.pdf', height=4, width=4)
corrplot(correlation_matrix, method = 'circle', addCoef.col = 'white', number.cex = 0.7)
dev.off()

# Batch 2 - same
batch_2_h2o <- tximport_function(dplyr::filter(batch2_conditions, condition == 'H2O control'), '/personal/path/to/batch2/kallisto/files')
# Log transform the gene abundances
abundances <- log(batch_2_h2o$abundance + 0.5) # using same log transformation we used for PCA
correlation_matrix <- cor(abundances)
pdf('corrplot/batch2.pdf', height=4, width=4)
corrplot(correlation_matrix, method = 'circle', addCoef.col = 'white', number.cex = 0.7)
dev.off()

# Batch 3 - same
batch_3_h2o <- tximport_function(dplyr::filter(batch3_conditions, condition == 'H2O control'), '/personal/path/to/batch3/kallisto/files')
# Log transform the gene abundances
abundances <- log(batch_3_h2o$abundance + 0.5) # using same log transformation we used for PCA
correlation_matrix <- cor(abundances)
pdf('corrplot/batch3.pdf', height=4, width=4)
corrplot(correlation_matrix, method = 'circle', addCoef.col = 'white', number.cex = 0.7)
dev.off()

# Batch 4 - DMSO controls on different plates
batch_4_h2o <- tximport_function(dplyr::filter(batch4_conditions, condition == 'DMSO control'), '/personal/path/to/batch4/kallisto/files')
# Log transform the gene abundances
abundances <- log(batch_4_h2o$abundance + 0.5) # using same log transformation we used for PCA
correlation_matrix <- cor(abundances)
pdf('corrplot/batch4.pdf', height=4, width=4)
corrplot(correlation_matrix, method = 'circle', addCoef.col = 'white', number.cex = 0.7)
dev.off()
```