---
title: "Untitled"
output: html_document
date: "2025-07-21"
---

Let's install Sleuth before performing PCA on our RNA-seq data, in order to get normalized, abundance-filtered, and log-transformed TPM values which can be used as input for PCA. Sleuth can provide these from its object.

```{r}
#install.packages("BiocManager")
#BiocManager::install("rhdf5")
#install.packages("devtools")
#devtools::install_github("pachterlab/sleuth")
```

Load necessary libraries.

```{r}
library("sleuth")
library(dplyr)
library(ggplot2)
library(tidyr)
library("readxl")
library(grafify)
```

For each of the four batches, set up a table with the following columns:
- Sample
- Condition
- Path (to Kallisto results)

Write a function to do this.

Based on:
https://pachterlab.github.io/sleuth_walkthroughs/trapnell/analysis.html

and:
https://pachterlab.github.io/sleuth_walkthroughs/pval_agg/analysis.html
(for the fact that we can point directly to h5 files)

```{r}
conditions_table <- function(batch_number, kallisto_dir) {
  # Read in master spreadsheet
  table <- read_excel("sample_info_transcriptomic_screen.xlsx")
  
  # Filter to batch of interest
  table <- dplyr::filter(table, Batch == batch_number)
  
  # Remove unnecessary columns
  table <- dplyr::select(table, `Sample name`, Treatment)
  table <- dplyr::rename(table, sample = `Sample name`, condition = Treatment)
  
  # Add path
  # Modify this code according to your file organization
  table$path <- file.path(kallisto_dir, paste0('kallisto_out_', sub("^[^_]*_", "", table$sample)), paste0(table$sample, "_abundance.h5"))
  
  # Ensure that condition is a factor
  table$condition <- factor(table$condition, levels=unique(table$condition))
  return(table)
}
```

Call the function for each batch.

```{r}
# Batch 1
conditions_1 <- conditions_table(1, '/personal/path/to/batch1/kallisto/files')
# the kallisto outputs for each sample should be within an independent subfolder of the path specified above

# Batch 2
conditions_2 <- conditions_table(2, '/personal/path/to/batch2/kallisto/files')
# the kallisto outputs for each sample should be within an independent subfolder of the path specified above

# Batch 3
conditions_3 <- conditions_table(3, '/personal/path/to/batch3/kallisto/files')
# the kallisto outputs for each sample should be within an independent subfolder of the path specified above

# Batch 4
conditions_4 <- conditions_table(4, '/personal/path/to/batch4/kallisto/files')
# the kallisto outputs for each sample should be within an independent subfolder of the path specified above
```

Helper functions for PCA - df_for_pca and pca_function.

```{r}
# get normalized, abundance-filtered, log-transformed TPM values, in correct format for PCA analysis

df_for_pca <- function(relevant_conditions_df) {
  #so <- sleuth_prep(relevant_conditions_df, extra_bootstrap_summary = TRUE)
  so <- sleuth_prep(relevant_conditions_df) # we don't need the extra bootstrap summary for PCA
  norm_kallisto <- kallisto_table(so, use_filtered = TRUE, normalized = TRUE)
  norm_kallisto <- select(norm_kallisto, target_id, sample, tpm)
  norm_kallisto$tpm <- log(norm_kallisto$tpm + 0.5)
  norm_kallisto <- norm_kallisto %>% pivot_wider(names_from = target_id, values_from = tpm) %>% tibble::column_to_rownames(., "sample")
  return(norm_kallisto)
}

# actually conduct PCA analysis based on the DF
pca_function <- function(pca_df) {
  return(prcomp(pca_df, scale = TRUE))
}
```

Wrapper function for doing + plotting the PCA, based on one conditions table.

```{r}
pca_plotting <- function(conditions_df, fig_title, treatment_levels, ht=3.5, wd=6) {
  
  pca_result <- pca_function(df_for_pca(conditions_df))
  
  pc1_var <- round(100*summary(pca_result)$importance[2,1])
  pc2_var <- round(100*summary(pca_result)$importance[2,2])
  pca_result <- as.data.frame(pca_result$x[, 1:2]) %>% tibble::rownames_to_column(., var='sample') %>% merge(., conditions_df, by = 'sample') %>% dplyr::rename(., Treatment=condition)
  
  # specify factor levels provided in function call
  pca_result$Treatment <- factor(pca_result$Treatment,
                                 levels = treatment_levels)
  
  ggplot(pca_result, aes(x=PC1, y=PC2, color=Treatment)) + geom_point() +
    labs(x = paste0('PC1 (', pc1_var, '% Variance Explained)'),
         y = paste0('PC2 (', pc2_var, '% Variance Explained)'),
         color = 'Treatment') + theme_bw() + 
  theme(panel.border = element_blank(), 
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(), 
        axis.line = element_line(colour = "black"), 
        axis.title=element_text(size=14, color='black'), 
        axis.text.x=element_text(size=12, color = "black"), 
        axis.text.y=element_text(size=12, color='black'),
        legend.text = element_text(size=12, color='black'),
        legend.title = element_text(size=14, color='black')) 
  
  ggsave(fig_title, height=ht, width=wd)
  
  return(pca_result)
}
```

Make the specific PCAs we need for the paper.

```{r}
# Batch 1 all DMSO samples
conditions_1_dmso <- dplyr::filter(conditions_1, condition %in% c("DMSO control", "Phenylacetylglutamine", "Cholate", "Indoleacetate"))
plotted_1_dmso <- pca_plotting(conditions_1_dmso, 'pca_plots/batch_1_dmso_samples.pdf',
             c("DMSO control", "Phenylacetylglutamine", "Cholate", "Indoleacetate"))

# Batch 1 all H2O samples
conditions_1_h2o <- dplyr::filter(conditions_1, !(condition %in% c("DMSO control", "Phenylacetylglutamine", "Cholate", "Indoleacetate")))
plotted_1_h2o <- pca_plotting(conditions_1_h2o, 'pca_plots/batch_1_h2o_samples.pdf', unique(conditions_1_h2o$condition))

# Batch 2 all samples
plotted_2 <- pca_plotting(conditions_2, 'pca_plots/batch_2_all_samples.pdf', unique(conditions_2$condition))

# Batch 3 all samples
plotted_3 <- pca_plotting(conditions_3, 'pca_plots/batch_3_all_samples.pdf', unique(conditions_3$condition))

# Batch 4 all DMSO samples
conditions_4_dmso <- dplyr::filter(conditions_4, !(condition %in% c("H2O control", "LPS control", "Urate", "Lysine")))
plotted_4_dmso <- pca_plotting(conditions_4_dmso, 'pca_plots/batch_4_dmso_samples.pdf', unique(conditions_4_dmso$condition))

# Batch 4 all H2O samples
conditions_4_h2o <- dplyr::filter(conditions_4, condition %in% c("H2O control", "LPS control", "Urate", "Lysine"))
plotted_4_h2o <- pca_plotting(conditions_4_h2o, 'pca_plots/batch_4_h2o_samples.pdf', unique(conditions_4_h2o$condition))
```