---
title: "Untitled"
output: html_document
date: "2025-08-11"
---

Load libraries.

```{r}
library(dplyr)
```

Read in the three intersection tables.

```{r}
darst <- read.csv("Darst_Chen_intersection.txt", sep = '\t')
menni <- read.csv("Menni_Chen_intersection.tsv", sep = '\t')
dunn <- read.csv("Dunn_Chen_intersection.tsv", sep = '\t')
```

Columns desired in the output: 1) Study name, 2) Metabolite name in original study, 3) Up or down with age, 4) Matched microbial metabolite, 5) HMDB Matched microbial metabolite, 6) CAS Matched microbial metabolite.

Reformat each table accordingly.

```{r}
# Darst
darst$study <- 'Darst'
darst$up_or_down <- ifelse(darst$B_age > 0, 'Up', 'Down')
darst <- dplyr::select(darst, study, BIOCHEMICAL_Darst, up_or_down, name_Chen, HMDB, cas_Chen)
darst$HMDB <- ifelse(darst$HMDB == 'HMDB0003374 (Darst); HMDB0000214 (Chen)',
                     'HMDB0000214', darst$HMDB)
colnames(darst) <- c("Study", "Original_Met_Name", "Up_or_down_with_age",
                     "Matched_Microbial_Met", "HMDB_Microbial_Met", "CAS_Microbial_Met")

# Menni
menni$study <- 'Menni'
menni$up_or_down <- ifelse(menni$Beta_age_Menni > 0, 'Up', 'Down')
menni <- dplyr::select(menni, study, Metabolite_Menni, up_or_down, Name_Chen, HMDB_Chen, CAS_Chen)
colnames(menni) <- c("Study", "Original_Met_Name", "Up_or_down_with_age",
                     "Matched_Microbial_Met", "HMDB_Microbial_Met", "CAS_Microbial_Met")

# Dunn
dunn$study <- 'Dunn'
dunn$up_or_down <- ifelse(dunn$Diff_Old_Minus_Young > 0, 'Up', 'Down')
dunn <- dplyr::select(dunn, study, Metabolite_Dunn, up_or_down, Name_Chen, HMDB_Chen, CAS_Chen)
colnames(dunn) <- c("Study", "Original_Met_Name", "Up_or_down_with_age",
                     "Matched_Microbial_Met", "HMDB_Microbial_Met", "CAS_Microbial_Met")
```

Combine all.

```{r}
combined <- rbind(darst, dunn, menni)
```

How many unique metabolites are there (look at HMDB_Microbial_Met column)?

```{r}
length(unique(combined$HMDB_Microbial_Met))
```

Output table.

```{r}
write.table(combined, 'Darst_Menni_Dunn_Chen_intersection.tsv', sep = '\t', quote = FALSE, row.names = FALSE)
```