---
title: "Untitled"
output: html_document
date: '2023-08-23'
---

Read in the Darst associations.

```{r}
library(readxl)
library(dplyr)
library(stringr)

darst_all_assoc <- read_excel("Darst_Table_S1.xlsx", sheet = "Results")
```

Filter to the metabolites that are significantly associated with age overall.

```{r}
darst_sig_assoc <- dplyr::filter(darst_all_assoc, p_age <= 0.05)
```

Filter out the unknown metabolites.

```{r}
darst_sig_assoc <- dplyr::filter(darst_sig_assoc, !(str_detect(darst_sig_assoc$BIOCHEMICAL, "^X - ")))
```

Keep only the relevant columns.

```{r}
darst_sig_assoc <- dplyr::select(darst_sig_assoc, BIOCHEMICAL, SUPER.PATHWAY, SUB.PATHWAY, CAS, PUBCHEM, HMDB, B_age, p_age)
```

Read in the list of microbial metabolites.

```{r}
microb_metab <- read_excel("Modified_Chen_Table_S3_Human_Metabolites_Tab.xlsx")

# for the identifier column, if it starts with 'PUBCHEMCID' then strip off that part
microb_metab <- microb_metab %>%
  mutate(identifier = str_replace(identifier, "^PUBCHEMCID", ""))
```

Filter Darst table to those features whose:
- HMDB ID is present in microb_metab$identifier, OR
- PUBCHEM ID is present in microb_metab$identifier, OR
- CAS is present in microb_metab$cas

Rbind the resulting filtered tables, and remove duplicate rows.

```{r}
# noticed that Darst sig assoc uses two fewer zeros for the HMDB ID - add those in
darst_sig_assoc <- darst_sig_assoc %>%
  mutate(HMDB = str_replace(HMDB, "^HMDB(\\d{5})$", "HMDB00\\1"))

darst_sig_assoc_hmdb <- dplyr::filter(darst_sig_assoc, HMDB %in% microb_metab$identifier)
darst_sig_assoc_pubchem <- dplyr::filter(darst_sig_assoc, PUBCHEM %in% microb_metab$identifier)
darst_sig_assoc_cas <- dplyr::filter(darst_sig_assoc, CAS %in% microb_metab$cas)

# rbind the HMDB and CAS tables (because pubchem yielded nothing), remove duplicates
darst_sig_assoc_matched_hmdb_or_cas <- rbind(darst_sig_assoc_hmdb, darst_sig_assoc_cas) %>% distinct()
```

That only added one metabolite!
Merge Darst with microb_metab based on the HMDB column, then manually add on the last metabolite that matched on CAS.

```{r}
metab_to_add <- darst_sig_assoc_cas$BIOCHEMICAL[which(!(darst_sig_assoc_cas$BIOCHEMICAL %in% darst_sig_assoc_hmdb$BIOCHEMICAL))]
# ornithine

darst_sig_assoc_hmdb <- merge(darst_sig_assoc, microb_metab, by.x='HMDB', by.y='identifier')
# reorder and rename columns
darst_sig_assoc_hmdb <- dplyr::select(darst_sig_assoc_hmdb, BIOCHEMICAL, name, SUPER.PATHWAY, SUB.PATHWAY, Source, HMDB, CAS, cas, B_age, p_age)
colnames(darst_sig_assoc_hmdb) <- c("BIOCHEMICAL_Darst", "name_Chen", "SUPER.PATHWAY_Darst",
                                    "SUB.PATHWAY_Darst", "Source_Chen", "HMDB", "CAS_Darst",
                                    "cas_Chen", "B_age", "p_age")

ornithine_row <- data.frame(BIOCHEMICAL_Darst = "ornithine", name_Chen = "Ornithine HCl",
                            SUPER.PATHWAY_Darst = "Amino Acid", SUB.PATHWAY_Darst = "Urea cycle; Arginine and Proline Metabolism",
                            Source_Chen = "Human/Food/Bacteria", HMDB = "HMDB0003374 (Darst); HMDB0000214 (Chen)",
                            CAS_Darst = "3184-13-2", cas_Chen = "3184-13-2",
                            B_age = 0.0016,
                            p_age = 1.88376e-05)

darst_sig_assoc_matched_hmdb_or_cas <- rbind(darst_sig_assoc_hmdb, ornithine_row)

write.table(darst_sig_assoc_matched_hmdb_or_cas, file='Darst_Chen_intersection.txt', sep='\t', quote=FALSE, row.names=FALSE)
```